---
title: "Bacterial score-clinical indicator correlation network"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: show
---

This analysis calculates a bacterial response score for each 16S sample using a
fixed baseline model, matches the scores to longitudinal clinical indicators by
patient-cycle identifier and timepoint, evaluates Spearman correlations, and
displays significant pooled and time-specific associations in a circular
network.

# 1. Analysis settings

```{r setup, message=FALSE, warning=FALSE}
rm(list = ls())

try(Sys.setlocale("LC_CTYPE", "en_US.UTF-8"), silent = TRUE)

suppressPackageStartupMessages({
  library(dplyr)
  library(openxlsx)
  library(ggplot2)
})

set.seed(20260712)

input_file <- "000.input16S.Rds"
model_file <- file.path(
  "07.outcomeM1_AUC",
  "02_B_cell_reconstitution_associated_score_model",
  "M0_bacteria_ResponseM1_score_model.rds"
)
clinical_file <- "000.clinical_indicators_deidentified.xlsx"

analysis_name <- "008.bacteriaScoreClinicalCorrelation"
output_dir <- analysis_name
figure_dir <- file.path(output_dir, "figures")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.path = file.path(output_dir, "knitr_figures", "")
)

network_time_levels <- c("BL", "D0", "D14", "M1", "M2", "M3", "M4", "M5", "M6")
pooled_fdr_threshold <- 0.10
timepoint_p_threshold <- 0.05
minimum_timepoint_n <- 8
```

# 2. Input data and bacterial scores

```{r read-inputs-and-score}
input_16s <- readRDS(input_file)
if (!all(c("dat_rela", "metainfo") %in% names(input_16s))) {
  stop("000.input16S.Rds must contain dat_rela and metainfo.")
}

dat_rela <- input_16s$dat_rela
metainfo <- input_16s$metainfo

required_metadata <- c(
  "Patient_ID", "CART_Cycle_ID", "Sample_Name", "Timepoint"
)
missing_metadata <- setdiff(required_metadata, colnames(metainfo))
if (length(missing_metadata) > 0L) {
  stop("Missing metadata columns: ", paste(missing_metadata, collapse = ", "))
}

model_information <- readRDS(model_file)
if (is.null(model_information$model)) {
  stop("The model RDS does not contain a fitted model.")
}

bacteria_model <- model_information$model
bacteria_pseudocount <- model_information$bacteria_pseudocount_M0
if (is.null(bacteria_pseudocount)) {
  bacteria_pseudocount <- model_information$bacteria_pseudo_M0
}
if (is.null(bacteria_pseudocount) || !is.finite(bacteria_pseudocount)) {
  stop("The model RDS does not contain a valid M0 bacterial pseudocount.")
}

as_numeric_matrix <- function(x) {
  x <- as.matrix(x)
  storage.mode(x) <- "numeric"
  x
}

predict_score <- function(model, new_x) {
  missing_features <- setdiff(model$features, colnames(new_x))
  if (length(missing_features) > 0L) {
    missing_matrix <- matrix(
      NA_real_,
      nrow = nrow(new_x),
      ncol = length(missing_features),
      dimnames = list(rownames(new_x), missing_features)
    )
    new_x <- cbind(new_x, missing_matrix)
  }

  new_x <- new_x[, model$features, drop = FALSE]
  new_z <- sweep(sweep(new_x, 2, model$center, "-"), 2, model$spread, "/")
  score <- rowMeans(sweep(new_z, 2, model$direction, "*"), na.rm = TRUE)
  score[!is.finite(score)] <- NA_real_
  score
}

score_metadata <- metainfo %>%
  mutate(
    Patient_ID = as.character(Patient_ID),
    CART_Cycle_ID = as.character(CART_Cycle_ID),
    Sample_Name = as.character(Sample_Name),
    time_label = as.character(Timepoint),
    patient_cycle_id = paste(Patient_ID, CART_Cycle_ID, sep = "_")
  ) %>%
  filter(time_label %in% network_time_levels)

if (anyDuplicated(score_metadata$Sample_Name)) {
  stop("Sample_Name must uniquely identify rows in metainfo.")
}

has_bacteria <- score_metadata$Sample_Name %in% colnames(dat_rela)
bacteria_score <- rep(NA_real_, nrow(score_metadata))

available_features <- intersect(bacteria_model$features, rownames(dat_rela))
if (length(available_features) == 0L) {
  stop("None of the model features are present in dat_rela.")
}

if (any(has_bacteria)) {
  bacteria_raw <- t(as_numeric_matrix(dat_rela[
    available_features,
    score_metadata$Sample_Name[has_bacteria],
    drop = FALSE
  ]))
  colnames(bacteria_raw) <- available_features
  bacteria_x <- log10(bacteria_raw + bacteria_pseudocount)
  bacteria_score[has_bacteria] <- predict_score(bacteria_model, bacteria_x)
}

score_values <- score_metadata %>%
  transmute(
    patient_cycle_id,
    time_label,
    Sample_Name,
    bacteria_score = bacteria_score,
    has_bacteria_score = !is.na(bacteria_score)
  )

sample_counts <- score_values %>%
  group_by(time_label) %>%
  summarise(
    n_sample = n(),
    n_bacteria_score = sum(has_bacteria_score),
    .groups = "drop"
  ) %>%
  mutate(time_label = factor(time_label, levels = network_time_levels)) %>%
  arrange(time_label) %>%
  mutate(time_label = as.character(time_label))

sample_counts
```

# 3. Clinical indicators

```{r read-clinical-indicators}
read_clinical_sheet_long <- function(file, sheet) {
  clinical_sheet <- openxlsx::read.xlsx(
    file,
    sheet = sheet,
    colNames = FALSE,
    check.names = FALSE
  )

  if (nrow(clinical_sheet) < 2L || ncol(clinical_sheet) < 2L) {
    return(data.frame())
  }

  patient_cycle_ids <- as.character(unlist(clinical_sheet[1, -1]))
  time_labels <- as.character(unlist(clinical_sheet[-1, 1]))
  value_matrix <- clinical_sheet[-1, -1, drop = FALSE]
  colnames(value_matrix) <- patient_cycle_ids

  clinical_rows <- lapply(seq_along(patient_cycle_ids), function(column_index) {
    data.frame(
      clinical_variable = sheet,
      patient_cycle_id = patient_cycle_ids[column_index],
      time_label = time_labels,
      clinical_value = suppressWarnings(as.numeric(unlist(value_matrix[, column_index]))),
      stringsAsFactors = FALSE
    )
  })

  do.call(rbind, clinical_rows)
}

clinical_sheets <- openxlsx::getSheetNames(clinical_file)
if (length(clinical_sheets) == 0L) {
  stop("No clinical indicator sheets were found.")
}

clinical_long <- bind_rows(lapply(
  clinical_sheets,
  function(sheet_name) read_clinical_sheet_long(clinical_file, sheet_name)
)) %>%
  mutate(
    patient_cycle_id = as.character(patient_cycle_id),
    time_label = as.character(time_label)
  ) %>%
  filter(
    !is.na(patient_cycle_id),
    patient_cycle_id != "",
    time_label %in% network_time_levels,
    !is.na(clinical_value)
  )

score_clinical_matched <- score_values %>%
  filter(!is.na(bacteria_score)) %>%
  inner_join(
    clinical_long,
    by = c("patient_cycle_id", "time_label"),
    relationship = "many-to-many"
  )

data.frame(
  clinical_indicator_count = length(unique(clinical_long$clinical_variable)),
  matched_rows = nrow(score_clinical_matched),
  matched_patient_cycles = length(unique(score_clinical_matched$patient_cycle_id))
)
```

# 4. Spearman correlations

```{r correlation-analysis}
safe_spearman <- function(data, x_column, y_column) {
  data <- data[!is.na(data[[x_column]]) & !is.na(data[[y_column]]), ]
  if (
    nrow(data) < 4L ||
    length(unique(data[[x_column]])) < 2L ||
    length(unique(data[[y_column]])) < 2L
  ) {
    return(data.frame(n = nrow(data), rho = NA_real_, P_value = NA_real_))
  }

  correlation_test <- suppressWarnings(cor.test(
    data[[x_column]],
    data[[y_column]],
    method = "spearman",
    exact = FALSE
  ))

  data.frame(
    n = nrow(data),
    rho = unname(correlation_test$estimate),
    P_value = correlation_test$p.value
  )
}

correlation_pooled <- bind_rows(lapply(
  split(score_clinical_matched, score_clinical_matched$clinical_variable),
  function(data_i) {
    result <- safe_spearman(data_i, "bacteria_score", "clinical_value")
    data.frame(
      analysis = "all_timepoints_pooled",
      score_type = "bacteria",
      clinical_variable = unique(data_i$clinical_variable)[1],
      time_label = "ALL",
      result,
      stringsAsFactors = FALSE
    )
  }
))

correlation_by_time <- bind_rows(lapply(
  split(
    score_clinical_matched,
    interaction(
      score_clinical_matched$clinical_variable,
      score_clinical_matched$time_label,
      drop = TRUE
    )
  ),
  function(data_i) {
    result <- safe_spearman(data_i, "bacteria_score", "clinical_value")
    data.frame(
      analysis = "by_timepoint",
      score_type = "bacteria",
      clinical_variable = unique(data_i$clinical_variable)[1],
      time_label = unique(data_i$time_label)[1],
      result,
      stringsAsFactors = FALSE
    )
  }
))

correlation_results <- bind_rows(correlation_pooled, correlation_by_time) %>%
  group_by(analysis) %>%
  mutate(FDR_global = p.adjust(P_value, method = "BH")) %>%
  ungroup() %>%
  group_by(analysis, score_type) %>%
  mutate(FDR_within_score = p.adjust(P_value, method = "BH")) %>%
  ungroup() %>%
  arrange(analysis, P_value)

safe_plot_label <- function(x) {
  x_utf8 <- enc2utf8(as.character(x))
  x_utf8[x_utf8 == "铁蛋白"] <- "Ferritin"
  label <- iconv(x_utf8, from = "UTF-8", to = "ASCII//TRANSLIT", sub = "_")
  label <- gsub("_+", "_", label)
  label <- gsub("^_|_$", "", label)
  ifelse(is.na(label) | label == "", "non_ascii_indicator", label)
}

network_edges_pooled <- correlation_results %>%
  filter(
    analysis == "all_timepoints_pooled",
    !is.na(FDR_global),
    FDR_global < pooled_fdr_threshold
  ) %>%
  mutate(
    clinical_variable_label = safe_plot_label(clinical_variable),
    edge_sign = ifelse(rho >= 0, "positive", "negative"),
    edge_label = sprintf(
      "rho=%.2f; P=%.3g; FDR=%.3g",
      rho,
      P_value,
      FDR_global
    )
  ) %>%
  arrange(desc(abs(rho)))

network_edges_by_time <- correlation_results %>%
  filter(
    analysis == "by_timepoint",
    !is.na(P_value),
    P_value < timepoint_p_threshold,
    n >= minimum_timepoint_n
  ) %>%
  mutate(
    clinical_variable_label = safe_plot_label(clinical_variable),
    edge_sign = ifelse(rho >= 0, "positive", "negative"),
    score_time_node = paste(score_type, time_label, sep = "_"),
    edge_label = sprintf("rho=%.2f; P=%.3g; n=%s", rho, P_value, n)
  ) %>%
  arrange(P_value)

circular_network_edges <- bind_rows(
  network_edges_pooled %>%
    mutate(
      source_node = "BacteiraOvellALL",
      source_type = "Overall",
      network_level = "overall_FDR_global_lt_0.10",
      source_time = "ALL"
    ),
  network_edges_by_time %>%
    mutate(
      source_node = paste0("Bacteria_", time_label),
      source_type = "Timepoint",
      network_level = "timepoint_raw_P_lt_0.05_n_ge_8",
      source_time = time_label
    )
) %>%
  transmute(
    source_node,
    target_node = clinical_variable_label,
    source_type,
    network_level,
    source_time,
    clinical_variable,
    clinical_variable_label,
    time_label,
    n,
    rho,
    P_value,
    FDR_global,
    FDR_within_score,
    edge_sign,
    edge_label
  ) %>%
  arrange(source_type, source_time, P_value)

data.frame(
  pooled_edges = nrow(network_edges_pooled),
  timepoint_edges = nrow(network_edges_by_time),
  total_edges = nrow(circular_network_edges)
)
```

# 5. Save analysis tables

```{r save-analysis-tables}
write_csv_utf8 <- function(data, file_name) {
  utils::write.csv(
    data,
    file.path(output_dir, file_name),
    row.names = FALSE,
    fileEncoding = "UTF-8"
  )
}

analysis_notes <- data.frame(
  item = c(
    "random_seed",
    "bacteria_model",
    "clinical_input",
    "score_transformation",
    "clinical_matching",
    "correlation_method",
    "pooled_edge_threshold",
    "timepoint_edge_threshold",
    "pooled_correlation_caution"
  ),
  description = c(
    "20260712",
    model_file,
    clinical_file,
    "log10(relative abundance + M0 pseudocount), followed by fixed M0 centering, scaling, and feature directions",
    "Patient-cycle identifier and timepoint",
    "Spearman correlation with exact = FALSE",
    "BH FDR_global < 0.10",
    "Raw P < 0.05 and n >= 8",
    "Pooled rows are exploratory because repeated measurements from the same patient-cycle are not independent."
  ),
  stringsAsFactors = FALSE
)

write_csv_utf8(analysis_notes, "00_analysis_notes.csv")
write_csv_utf8(score_values, "01_bacteria_scores.csv")
write_csv_utf8(sample_counts, "02_sample_counts.csv")
write_csv_utf8(clinical_long, "03_clinical_indicators_long.csv")
write_csv_utf8(score_clinical_matched, "04_score_clinical_matched_long.csv")
write_csv_utf8(correlation_results, "05_spearman_correlations.csv")
write_csv_utf8(circular_network_edges, "06_circular_network_edges.csv")
```

# 6. Circular correlation network

```{r circular-network, fig.width=10.2, fig.height=8.4}
plot_bacteria_circular_network <- function(edges) {
  if (nrow(edges) == 0L) {
    stop("No edges passed the specified network thresholds.")
  }

  source_nodes <- edges %>%
    distinct(node = source_node) %>%
    mutate(node_type = "Bacteria score")

  clinical_nodes <- edges %>%
    distinct(node = target_node) %>%
    mutate(node_type = "Clinical indicator")

  node_degree <- c(edges$source_node, edges$target_node) %>%
    table() %>%
    as.data.frame(stringsAsFactors = FALSE)
  colnames(node_degree) <- c("node", "degree")
  node_degree$degree <- as.numeric(node_degree$degree)

  source_order <- source_nodes %>%
    mutate(
      source_rank = case_when(
        node == "BacteiraOvellALL" ~ 0,
        TRUE ~ match(sub("^Bacteria_", "", node), network_time_levels)
      )
    ) %>%
    arrange(source_rank) %>%
    pull(node)

  clinical_order <- edges %>%
    group_by(target_node) %>%
    summarise(
      degree = n(),
      min_p = min(P_value, na.rm = TRUE),
      max_abs_rho = max(abs(rho), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      min_p = ifelse(is.finite(min_p), min_p, Inf),
      max_abs_rho = ifelse(is.finite(max_abs_rho), max_abs_rho, 0)
    ) %>%
    arrange(desc(degree), min_p, desc(max_abs_rho)) %>%
    pull(target_node)

  node_order <- c(source_order, clinical_order)
  node_plot <- bind_rows(source_nodes, clinical_nodes) %>%
    distinct(node, .keep_all = TRUE) %>%
    left_join(node_degree, by = "node") %>%
    mutate(
      degree = ifelse(is.na(degree), 0, degree),
      node = factor(node, levels = node_order),
      node_index = as.numeric(node),
      angle = pi / 2 - 2 * pi * (node_index - 1) / length(node_order),
      x = cos(angle),
      y = sin(angle),
      label_x = 1.16 * cos(angle),
      label_y = 1.16 * sin(angle),
      label_hjust = ifelse(label_x > 0.02, 0, ifelse(label_x < -0.02, 1, 0.5)),
      label_vjust = ifelse(label_y > 0.92, 0, ifelse(label_y < -0.92, 1, 0.5))
    )

  edge_plot <- edges %>%
    left_join(
      node_plot %>% dplyr::select(source_node = node, x_source = x, y_source = y),
      by = "source_node"
    ) %>%
    left_join(
      node_plot %>% dplyr::select(target_node = node, x_target = x, y_target = y),
      by = "target_node"
    ) %>%
    mutate(
      abs_rho = abs(rho),
      edge_sign = ifelse(rho >= 0, "positive", "negative"),
      edge_alpha = ifelse(source_type == "Overall", 0.90, 0.58)
    )

  ggplot() +
    geom_segment(
      data = edge_plot,
      aes(
        x = x_source,
        y = y_source,
        xend = x_target,
        yend = y_target,
        color = edge_sign,
        linewidth = abs_rho,
        alpha = edge_alpha
      ),
      lineend = "round"
    ) +
    geom_point(
      data = node_plot,
      aes(x = x, y = y, size = degree, fill = node_type),
      shape = 21,
      color = "grey20",
      stroke = 0.45
    ) +
    geom_text(
      data = node_plot,
      aes(
        x = label_x,
        y = label_y,
        label = as.character(node),
        hjust = label_hjust,
        vjust = label_vjust
      ),
      size = 3.2,
      lineheight = 0.95
    ) +
    scale_color_manual(
      values = c(positive = "#B2182B", negative = "#2166AC"),
      name = "Correlation"
    ) +
    scale_fill_manual(
      values = c("Bacteria score" = "#FDE68A", "Clinical indicator" = "#DBEAFE"),
      name = "Node"
    ) +
    scale_size_continuous(range = c(3.8, 12), name = "Degree") +
    scale_linewidth(range = c(0.35, 1.9), name = "abs(rho)") +
    scale_alpha_identity() +
    coord_equal(
      xlim = c(-1.45, 1.45),
      ylim = c(-1.35, 1.35),
      clip = "off"
    ) +
    theme_void(base_size = 11) +
    theme(
      legend.position = "right",
      legend.box = "vertical",
      plot.margin = margin(16, 24, 16, 24),
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    ) +
    guides(
      fill = guide_legend(order = 1),
      size = guide_legend(order = 2),
      linewidth = guide_legend(order = 3),
      color = guide_legend(order = 4)
    ) +
    labs(
      title = "Bacteria score-clinical correlation circular network",
      subtitle = paste0(
        "BacteiraOvellALL: pooled bacteria correlations with FDR<0.10; ",
        "Bacteria_time: timepoint correlations with raw P<0.05 and n>=8"
      )
    )
}

circular_network_plot <- plot_bacteria_circular_network(circular_network_edges)
print(circular_network_plot)

ggsave(
  file.path(
    figure_dir,
    "Figure_10_bacteria_score_clinical_correlation_circular_network.pdf"
  ),
  circular_network_plot,
  width = 10.2,
  height = 8.4
)
```

# 7. Session information

```{r session-information}
sessionInfo()
```
