---
title: "Metagenomic pathway and KO functional analysis"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    self_contained: true
---

This workflow evaluates longitudinal changes in metagenomic pathway and KEGG
ortholog abundance, performs pathway-level differential analysis and gene set
enrichment analysis, and compares KO enrichment patterns between clinical
outcome groups at each sampling timepoint.

```{r setup, include=FALSE}
options(stringsAsFactors = FALSE)

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(limma)
  library(ComplexHeatmap)
  library(circlize)
  library(clusterProfiler)
  library(enrichplot)
  library(openxlsx)
})

input_file <- "000.input16S.Rds"
pathway_file <- file.path("009.pathway", "000.pathabundance.deidentified.Rds")
ko_file <- file.path("009.pathway", "000.KO_unstratified.deidentified.Rds")
kegg_file <- file.path("009.pathway", "000.KEGG_KO_pathway_annotation.Rds")

analysis_name <- "009.pathway"
output_dir <- analysis_name
figure_dir <- file.path(output_dir, "figures")
table_dir <- file.path(output_dir, "tables")
outcome_dir <- file.path(output_dir, "outcome_KO_GSEA")

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(outcome_dir, recursive = TRUE, showWarnings = FALSE)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.path = file.path(output_dir, "knitr_figures", "")
)

time_levels <- c(
  "001_BL", "002_D0", "003_D14", "004_M1",
  "005_M2", "006_M3", "009_M6"
)
comparison_names <- c(
  "D0_vs_BL", "D14_vs_BL", "M1_vs_BL",
  "M2_vs_BL", "M3_vs_BL", "M6_vs_BL"
)
pseudo_count <- 1e-6
prevalence_fraction <- 0.20
```

# Input preparation

```{r read-inputs}
required_files <- c(input_file, pathway_file, ko_file, kegg_file)
missing_files <- required_files[!file.exists(required_files)]
if (length(missing_files) > 0L) {
  stop("Missing input file(s): ", paste(missing_files, collapse = ", "))
}

input16S <- readRDS(input_file)
if (!is.list(input16S) || !all(c("dat_rela", "metainfo") %in% names(input16S))) {
  stop("000.input16S.Rds must contain dat_rela and metainfo.")
}

metainfo <- input16S$metainfo
required_metadata <- c("Sample_Name", "timeIDcontiue", "outcome_M1", "outcome_M3")
missing_metadata <- setdiff(required_metadata, names(metainfo))
if (length(missing_metadata) > 0L) {
  stop("Missing metadata column(s): ", paste(missing_metadata, collapse = ", "))
}
if (anyDuplicated(metainfo$Sample_Name)) {
  stop("metainfo$Sample_Name must be unique.")
}

pathway_raw <- as.matrix(readRDS(pathway_file))
ko_raw <- as.matrix(readRDS(ko_file))
storage.mode(pathway_raw) <- "numeric"
storage.mode(ko_raw) <- "numeric"

if (anyDuplicated(colnames(pathway_raw)) || anyDuplicated(colnames(ko_raw))) {
  stop("The abundance matrices contain duplicated Sample_Name columns.")
}
if (!all(colnames(pathway_raw) %in% metainfo$Sample_Name) ||
    !all(colnames(ko_raw) %in% metainfo$Sample_Name)) {
  stop("Every abundance column must match metainfo$Sample_Name.")
}

metadata_for <- function(sample_names, allowed_times = time_levels) {
  ans <- metainfo[match(sample_names, metainfo$Sample_Name), , drop = FALSE]
  keep <- !is.na(ans$timeIDcontiue) & ans$timeIDcontiue %in% allowed_times
  ans <- ans[keep, , drop = FALSE]
  rownames(ans) <- ans$Sample_Name
  ans
}

to_cpm <- function(x) {
  totals <- colSums(x, na.rm = TRUE)
  if (any(!is.finite(totals) | totals <= 0)) {
    stop("All retained samples must have a positive finite library total.")
  }
  sweep(x, 2, totals, "/") * 1e6
}

fit_time_contrasts <- function(y, metadata) {
  metadata <- metadata[colnames(y), , drop = FALSE]
  group <- factor(metadata$timeIDcontiue, levels = time_levels)
  design <- model.matrix(~ 0 + group)
  colnames(design) <- make.names(time_levels)

  fit <- eBayes(lmFit(y, design), trend = TRUE)
  contrast_matrix <- makeContrasts(
    D0_vs_BL  = X002_D0  - X001_BL,
    D14_vs_BL = X003_D14 - X001_BL,
    M1_vs_BL  = X004_M1  - X001_BL,
    M2_vs_BL  = X005_M2  - X001_BL,
    M3_vs_BL  = X006_M3  - X001_BL,
    M6_vs_BL  = X009_M6  - X001_BL,
    levels = design
  )
  fit2 <- eBayes(contrasts.fit(fit, contrast_matrix), trend = TRUE)

  results <- lapply(comparison_names, function(comparison) {
    tab <- topTable(
      fit2, coef = comparison, n = Inf,
      adjust.method = "BH", sort.by = "P"
    )
    tab$feature <- rownames(tab)
    tab$type <- comparison
    rownames(tab) <- NULL
    tab
  })
  names(results) <- comparison_names
  results
}
```

# Longitudinal pathway analysis

```{r pathway-differential-analysis}
pathway_metadata <- metadata_for(colnames(pathway_raw))
pathway_dat <- pathway_raw[, rownames(pathway_metadata), drop = FALSE]
pathway_dat <- pathway_dat[
  !grepl("UNMAPPED|UNINTEGRATED|unclassified", rownames(pathway_dat), ignore.case = TRUE) &
    !grepl("g__", rownames(pathway_dat)),
  , drop = FALSE
]
pathway_dat <- pathway_dat[rowSums(pathway_dat > 0) > ncol(pathway_dat) * prevalence_fraction, , drop = FALSE]
pathway_cpm <- to_cpm(pathway_dat)
pathway_clr <- apply(
  pathway_cpm + pseudo_count,
  2,
  function(x) log(x) - mean(log(x))
)

pathway_results <- fit_time_contrasts(pathway_clr, pathway_metadata)
pathway_results_long <- bind_rows(pathway_results)
pathway_results_long <- pathway_results_long %>%
  rename(pathway = feature)

saveRDS(
  list(
    differential_results = pathway_results,
    cpm = pathway_cpm,
    metadata = pathway_metadata[, c("Sample_Name", "timeIDcontiue"), drop = FALSE]
  ),
  file.path(table_dir, "longitudinal_pathway_limma_results.rds")
)
write.csv(
  pathway_results_long,
  file.path(table_dir, "longitudinal_pathway_limma_results.csv"),
  row.names = FALSE
)

pathway_sample_counts <- pathway_metadata %>%
  count(timeIDcontiue, name = "sample_count")
write.csv(
  pathway_sample_counts,
  file.path(table_dir, "longitudinal_pathway_sample_counts.csv"),
  row.names = FALSE
)

knitr::kable(pathway_sample_counts)
```

## Combined abundance and differential heatmaps

```{r pathway-heatmaps}
raw_p_rank <- pathway_results_long %>%
  filter(P.Value < 0.05) %>%
  count(pathway, name = "raw_p_count") %>%
  arrange(desc(raw_p_count)) %>%
  pull(pathway)

adjusted_significant <- pathway_results_long %>%
  filter(adj.P.Val < 0.05) %>%
  pull(pathway) %>%
  unique()

recurrent_raw_p <- pathway_results_long %>%
  filter(P.Value < 0.05) %>%
  count(pathway) %>%
  filter(n > 3) %>%
  pull(pathway)

selected_pathways <- raw_p_rank[
  raw_p_rank %in% union(adjusted_significant, recurrent_raw_p)
]
if (length(selected_pathways) == 0L) {
  stop("No pathway met the heatmap selection criteria.")
}

top15_pathways <- head(selected_pathways, 15)
remaining_pathways <- if (length(selected_pathways) > 15) {
  selected_pathways[-seq_len(15)]
} else {
  selected_pathways
}

get_star <- function(p) {
  ifelse(
    is.na(p), "",
    ifelse(
      p < 0.001, "***",
      ifelse(p < 0.01, "**", ifelse(p < 0.05, "*", ifelse(p < 0.1, ".", "")))
    )
  )
}

draw_pathway_heatmap <- function(selected, filename, height) {
  abundance <- pathway_cpm[selected, rownames(pathway_metadata), drop = FALSE]
  abundance_scaled <- t(scale(t(log2(abundance + 1))))
  abundance_scaled[!is.finite(abundance_scaled)] <- 0

  differential_subset <- pathway_results_long %>%
    filter(pathway %in% selected)

  logfc_mat <- differential_subset %>%
    select(pathway, type, logFC) %>%
    pivot_wider(names_from = type, values_from = logFC, values_fill = 0) %>%
    column_to_rownames("pathway") %>%
    as.matrix()
  logfc_mat <- logfc_mat[selected, comparison_names, drop = FALSE]

  pval_mat <- differential_subset %>%
    select(pathway, type, P.Value) %>%
    pivot_wider(names_from = type, values_from = P.Value, values_fill = 1) %>%
    column_to_rownames("pathway") %>%
    as.matrix()
  pval_mat <- pval_mat[selected, comparison_names, drop = FALSE]

  padj_mat <- differential_subset %>%
    select(pathway, type, adj.P.Val) %>%
    pivot_wider(names_from = type, values_from = adj.P.Val, values_fill = 1) %>%
    column_to_rownames("pathway") %>%
    as.matrix()
  padj_mat <- padj_mat[selected, comparison_names, drop = FALSE]
  star_mat <- matrix(get_star(pval_mat), nrow = nrow(pval_mat), dimnames = dimnames(pval_mat))

  group_info <- factor(pathway_metadata$timeIDcontiue, levels = time_levels)
  group_colors <- c(
    "001_BL" = "#F2E778", "002_D0" = "#F5B183", "003_D14" = "#E99568",
    "004_M1" = "#EC83C3", "005_M2" = "#F3D768", "006_M3" = "#82DFC6",
    "009_M6" = "#E77AE2"
  )
  top_annotation <- HeatmapAnnotation(
    Group = group_info,
    col = list(Group = group_colors)
  )

  max_logfc <- max(abs(logfc_mat), na.rm = TRUE)
  if (!is.finite(max_logfc) || max_logfc == 0) max_logfc <- 1

  ht_abundance <- Heatmap(
    abundance_scaled,
    name = "Z-score",
    col = colorRamp2(c(-2, 0, 2), c("#3182BD", "white", "#D73027")),
    top_annotation = top_annotation,
    column_split = group_info,
    column_gap = grid::unit(2, "mm"),
    cluster_columns = FALSE,
    cluster_rows = FALSE,
    show_column_names = FALSE,
    show_row_names = TRUE,
    row_names_side = "left",
    row_names_gp = grid::gpar(fontsize = 8),
    width = grid::unit(15, "cm"),
    column_title = "Abundance (Per Sample)"
  )

  ht_logfc <- Heatmap(
    logfc_mat,
    name = "logFC",
    col = colorRamp2(
      c(-max_logfc, 0, max_logfc),
      c("blue", "white", "red")
    ),
    cluster_columns = FALSE,
    cluster_rows = FALSE,
    show_row_names = FALSE,
    column_names_gp = grid::gpar(fontsize = 10),
    width = grid::unit(4, "cm"),
    column_title = "logFC & Significance",
    cell_fun = function(j, i, x, y, width, height, fill) {
      if (!is.na(padj_mat[i, j]) && padj_mat[i, j] < 0.05) {
        grid::grid.rect(
          x = x, y = y, width = width, height = height,
          gp = grid::gpar(col = "black", fill = NA, lwd = 1.5)
        )
      }
      if (star_mat[i, j] != "") {
        grid::grid.text(
          star_mat[i, j], x, y, vjust = 0.7,
          gp = grid::gpar(fontsize = 12, col = "black")
        )
      }
    }
  )

  grDevices::pdf(filename, width = 14.29, height = height)
  draw(
    ht_abundance + ht_logfc,
    main_heatmap = "Z-score",
    padding = grid::unit(c(2, 2, 2, 2), "cm")
  )
  grDevices::dev.off()
}

draw_pathway_heatmap(
  top15_pathways,
  file.path(figure_dir, "004.heatmap_combined.pdf"),
  height = 5.08
)
draw_pathway_heatmap(
  remaining_pathways,
  file.path(figure_dir, "004.heatmap_combined.excluding_top15.pdf"),
  height = 5.08 * 3
)

write.csv(
  data.frame(
    pathway = selected_pathways,
    heatmap_set = ifelse(selected_pathways %in% top15_pathways, "top15", "excluding_top15")
  ),
  file.path(table_dir, "pathway_heatmap_features.csv"),
  row.names = FALSE
)
```

## Pathway volcano plot

```{r pathway-volcano}
volcano_data <- pathway_results_long %>%
  mutate(
    color_group = case_when(
      P.Value < 0.05 & logFC > 0 ~ "Up",
      P.Value < 0.05 & logFC < 0 ~ "Down",
      TRUE ~ "NS"
    ),
    shape_group = ifelse(adj.P.Val < 0.05, "adjSig", "nonAdj"),
    type = factor(type, levels = comparison_names)
  )

volcano_plot <- ggplot(
  volcano_data,
  aes(
    x = logFC, y = -log10(P.Value),
    size = shape_group, color = color_group, shape = shape_group
  )
) +
  geom_point(alpha = 0.8) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("Up" = "#D73027", "Down" = "#4575B4", "NS" = "grey70")) +
  scale_shape_manual(values = c("adjSig" = 1, "nonAdj" = 16)) +
  scale_size_manual(values = c("adjSig" = 4, "nonAdj" = 1)) +
  facet_wrap(~ type, nrow = 2) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    legend.title = element_blank(),
    plot.title = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_rect(fill = NA, colour = "grey87")
  )

ggsave(
  file.path(figure_dir, "004.volcano.pdf"),
  plot = volcano_plot,
  width = 11.15,
  height = 4.45
)
print(volcano_plot)
```

# Longitudinal KO analysis and GSEA

```{r longitudinal-ko-differential}
ko_metadata <- metadata_for(colnames(ko_raw))
ko_dat <- ko_raw[, rownames(ko_metadata), drop = FALSE]
ko_dat <- ko_dat[!grepl("g__", rownames(ko_dat)), , drop = FALSE]
ko_cpm <- to_cpm(ko_dat)
ko_cpm <- ko_cpm[
  rowSums(ko_cpm > 0) > ncol(ko_cpm) * prevalence_fraction,
  , drop = FALSE
]
ko_log2 <- log2(ko_cpm + pseudo_count)

ko_results <- fit_time_contrasts(ko_log2, ko_metadata)
ko_results_long <- bind_rows(ko_results) %>%
  rename(KO = feature)

saveRDS(
  list(
    differential_results = ko_results,
    cpm = ko_cpm,
    metadata = ko_metadata[, c("Sample_Name", "timeIDcontiue"), drop = FALSE]
  ),
  file.path(table_dir, "longitudinal_KO_limma_results.rds")
)
write.csv(
  ko_results_long,
  file.path(table_dir, "longitudinal_KO_limma_results.csv"),
  row.names = FALSE
)
```

```{r longitudinal-ko-gsea}
kegg_dat <- readRDS(kegg_file)
required_kegg_columns <- c("A", "C_code", "C_name", "K")
if (!all(required_kegg_columns %in% names(kegg_dat))) {
  stop("The KEGG annotation file is missing required columns.")
}

run_longitudinal_gsea <- function(category) {
  annotation <- kegg_dat %>%
    filter(A != "A09160 Human Diseases", A == category)
  term2gene <- annotation %>%
    select(term = C_code, gene = K) %>%
    distinct()
  term2name <- annotation %>%
    select(term = C_code, name = C_name) %>%
    distinct()

  ans <- lapply(comparison_names, function(comparison) {
    tab <- ko_results[[comparison]]
    gene_list <- tab$logFC
    names(gene_list) <- tab$feature
    gene_list <- sort(gene_list[is.finite(gene_list)], decreasing = TRUE)
    term2gene_use <- term2gene[term2gene$gene %in% names(gene_list), , drop = FALSE]

    GSEA(
      geneList = gene_list,
      TERM2GENE = term2gene_use,
      TERM2NAME = term2name,
      pvalueCutoff = 1,
      verbose = FALSE
    )
  })
  names(ans) <- comparison_names
  ans
}

metabolism_gsea <- run_longitudinal_gsea("A09100 Metabolism")
brite_gsea <- run_longitudinal_gsea("A09180 Brite Hierarchies")

saveRDS(metabolism_gsea, file.path(table_dir, "longitudinal_KO_GSEA_metabolism.rds"))
saveRDS(brite_gsea, file.path(table_dir, "longitudinal_KO_GSEA_brite.rds"))

gsea_to_table <- function(x) {
  bind_rows(lapply(names(x), function(comparison) {
    tab <- as.data.frame(x[[comparison]])
    tab$Comparison <- comparison
    tab
  }))
}
write.csv(
  gsea_to_table(metabolism_gsea),
  file.path(table_dir, "longitudinal_KO_GSEA_metabolism.csv"),
  row.names = FALSE
)
write.csv(
  gsea_to_table(brite_gsea),
  file.path(table_dir, "longitudinal_KO_GSEA_brite.csv"),
  row.names = FALSE
)
```

## Selected GSEA enrichment curves

```{r selected-gsea-plots}
require_gene_sets <- function(gsea_object, ids, label) {
  available <- as.data.frame(gsea_object)$ID
  missing <- setdiff(ids, available)
  if (length(missing) > 0L) {
    stop(label, " is missing GSEA set(s): ", paste(missing, collapse = ", "))
  }
}

metabolism_ids <- c("ko00121", "ko00640", "ko00920")
require_gene_sets(metabolism_gsea[["M1_vs_BL"]], metabolism_ids, "M1_vs_BL metabolism")
require_gene_sets(metabolism_gsea[["D14_vs_BL"]], metabolism_ids, "D14_vs_BL metabolism")

gsea_m1 <- gseaplot2(
  metabolism_gsea[["M1_vs_BL"]],
  geneSetID = metabolism_ids,
  color = c("#E41749", "#007A87", "#FFB400"),
  subplots = c(1, 2, 3),
  pvalue_table = TRUE
)
ggsave(
  file.path(figure_dir, "005.GSEA.M1_vs_BL.pdf"),
  plot = gsea_m1, width = 6.98, height = 3.11
)

gsea_d14 <- gseaplot2(
  metabolism_gsea[["D14_vs_BL"]],
  geneSetID = metabolism_ids,
  color = c("#E41749", "#007A87", "#FFB400"),
  subplots = c(1, 2, 3),
  pvalue_table = TRUE
)
ggsave(
  file.path(figure_dir, "005.GSEA.D14_vs_BL.pdf"),
  plot = gsea_d14, width = 6.98, height = 3.11
)

amr_id <- "ko01504"
require_gene_sets(brite_gsea[["M1_vs_BL"]], amr_id, "M1_vs_BL Brite")
require_gene_sets(brite_gsea[["M3_vs_BL"]], amr_id, "M3_vs_BL Brite")

gsea_m1_amr <- gseaplot2(
  brite_gsea[["M1_vs_BL"]],
  geneSetID = amr_id,
  color = "#FCCDE5",
  subplots = c(1, 2, 3),
  pvalue_table = TRUE
)
ggsave(
  file.path(figure_dir, "005.GSEA.M1_vs_BL.antimicrobial_resistance.pdf"),
  plot = gsea_m1_amr, width = 11.53, height = 3.23
)

gsea_m3_amr <- gseaplot2(
  brite_gsea[["M3_vs_BL"]],
  geneSetID = amr_id,
  color = "#FCCDE5",
  subplots = c(1, 2, 3),
  pvalue_table = TRUE
)
ggsave(
  file.path(figure_dir, "005.GSEA.M3_vs_BL.antimicrobial_resistance.pdf"),
  plot = gsea_m3_amr, width = 11.53, height = 3.23
)

print(gsea_d14)
```

# KO comparisons by clinical outcome

The outcome analyses use the same comparison direction as the source analysis:
`later - early`. Positive log-fold changes and positive NES values therefore
indicate higher values in the later-response group.

```{r outcome-analysis-functions}
outcome_time_levels <- time_levels[time_levels != "009_M6"]
outcome_definitions <- list(
  Result_M1 = list(
    metadata_column = "outcome_M1",
    group_map = c("Fast_responders" = "early", "Delayed_responders" = "later")
  ),
  Result_M3 = list(
    metadata_column = "outcome_M3",
    group_map = c("Responder" = "early", "Non-responder" = "later")
  )
)
min_group_n <- 2

empty_outcome_result <- function() {
  data.frame(
    logFC = numeric(), AveExpr = numeric(), t = numeric(),
    P.Value = numeric(), adj.P.Val = numeric(), B = numeric(),
    Mean_early = numeric(), Mean_later = numeric(),
    KO = character(), timeID = character(), outcome = character(),
    comparison = character(), stringsAsFactors = FALSE
  )
}

fit_outcome_by_time <- function(y, metadata, outcome_name, outcome_column, group_map) {
  mapped_group <- unname(group_map[as.character(metadata[[outcome_column]])])
  results <- list()
  sample_counts <- list()

  for (timepoint in outcome_time_levels) {
    keep <- metadata$timeIDcontiue == timepoint &
      !is.na(mapped_group) & mapped_group %in% c("early", "later")
    group <- factor(mapped_group[keep], levels = c("early", "later"))
    group_n <- table(group)

    sample_counts[[timepoint]] <- data.frame(
      outcome = outcome_name,
      timeID = timepoint,
      group = names(group_n),
      n = as.integer(group_n)
    )

    if (length(group) == 0L || any(group_n < min_group_n)) {
      results[[timepoint]] <- empty_outcome_result()
      next
    }

    y_time <- y[, keep, drop = FALSE]
    design <- model.matrix(~ 0 + group)
    colnames(design) <- levels(group)
    fit <- eBayes(lmFit(y_time, design), trend = TRUE)
    contrast_matrix <- makeContrasts(later - early, levels = design)
    fit2 <- eBayes(contrasts.fit(fit, contrast_matrix), trend = TRUE)

    tab <- topTable(fit2, coef = 1, n = Inf, adjust.method = "BH", sort.by = "P")
    tab$Mean_early <- fit$coefficients[rownames(tab), "early"]
    tab$Mean_later <- fit$coefficients[rownames(tab), "later"]
    tab$KO <- rownames(tab)
    tab$timeID <- timepoint
    tab$outcome <- outcome_name
    tab$comparison <- "later_vs_early"
    rownames(tab) <- NULL
    results[[timepoint]] <- tab
  }

  list(results = results, sample_counts = bind_rows(sample_counts))
}

run_outcome_gsea <- function(differential_results) {
  annotation <- kegg_dat %>%
    filter(A != "A09160 Human Diseases", A == "A09100 Metabolism")
  term2gene <- annotation %>%
    select(term = C_code, gene = K) %>%
    distinct()
  term2name <- annotation %>%
    select(term = C_code, name = C_name) %>%
    distinct()

  ans <- lapply(names(differential_results), function(timepoint) {
    tab <- differential_results[[timepoint]]
    if (nrow(tab) == 0L) return(NULL)
    gene_list <- tab$logFC
    names(gene_list) <- tab$KO
    gene_list <- sort(
      gene_list[is.finite(gene_list) & !duplicated(names(gene_list))],
      decreasing = TRUE
    )
    term2gene_use <- term2gene[term2gene$gene %in% names(gene_list), , drop = FALSE]
    GSEA(
      geneList = gene_list,
      TERM2GENE = term2gene_use,
      TERM2NAME = term2name,
      pvalueCutoff = 1,
      minGSSize = 5,
      maxGSSize = 500,
      verbose = FALSE
    )
  })
  names(ans) <- names(differential_results)
  ans
}

plot_outcome_heatmap <- function(
  enrichment_table,
  outcome_name,
  all_significant = FALSE
) {
  p_cutoff <- 0.05
  top_n <- 15
  enrichment_table <- enrichment_table %>%
    filter(is.finite(NES), !is.na(p.adjust)) %>%
    distinct(timeID, ID, .keep_all = TRUE)

  significant_ids <- enrichment_table %>%
    filter(p.adjust <= p_cutoff) %>%
    pull(ID) %>%
    unique()

  if (all_significant) {
    selected_ids <- significant_ids
  } else {
    source_table <- if (length(significant_ids) > 0L) {
      enrichment_table %>% filter(ID %in% significant_ids)
    } else {
      enrichment_table
    }
    selected_ids <- source_table %>%
      group_by(timeID) %>%
      slice_min(order_by = p.adjust, n = top_n, with_ties = FALSE) %>%
      ungroup() %>%
      pull(ID) %>%
      unique()
  }

  if (length(selected_ids) == 0L) {
    plot <- ggplot() +
      annotate("text", x = 1, y = 1, label = "No GSEA result") +
      theme_void()
    return(list(plot = plot, height = 6))
  }

  pathway_names <- enrichment_table %>%
    filter(ID %in% selected_ids) %>%
    mutate(
      term_label = ifelse(
        is.na(Description) | Description == "",
        ID,
        paste0(Description, " [", ID, "]")
      )
    ) %>%
    group_by(ID) %>%
    summarise(term_label = first(term_label), .groups = "drop")

  heatmap_data <- expand.grid(
    timeID = outcome_time_levels,
    ID = selected_ids,
    stringsAsFactors = FALSE
  ) %>%
    left_join(
      enrichment_table %>%
        filter(ID %in% selected_ids) %>%
        select(timeID, ID, NES, p.adjust),
      by = c("timeID", "ID")
    ) %>%
    left_join(pathway_names, by = "ID") %>%
    mutate(
      significance = case_when(
        !is.na(p.adjust) & p.adjust < 0.001 ~ "***",
        !is.na(p.adjust) & p.adjust < 0.01 ~ "**",
        !is.na(p.adjust) & p.adjust <= p_cutoff ~ "*",
        TRUE ~ ""
      ),
      timeID = factor(timeID, levels = outcome_time_levels)
    )

  term_order <- enrichment_table %>%
    filter(ID %in% selected_ids) %>%
    group_by(ID) %>%
    summarise(best_p = min(p.adjust, na.rm = TRUE), .groups = "drop") %>%
    left_join(pathway_names, by = "ID") %>%
    arrange(desc(best_p)) %>%
    pull(term_label)
  heatmap_data$term_label <- factor(heatmap_data$term_label, levels = term_order)

  subtitle <- if (all_significant) {
    "Pathways significant at one or more timepoints; * adjusted P <= 0.05"
  } else {
    "Top 15 pathways per timepoint; * adjusted P < 0.05"
  }

  plot <- ggplot(heatmap_data, aes(x = timeID, y = term_label, fill = NES)) +
    geom_tile(color = "white", linewidth = 0.25) +
    geom_text(aes(label = significance), size = 3) +
    scale_fill_gradient2(
      low = "#2166AC", mid = "white", high = "#B2182B",
      midpoint = 0, na.value = "grey90", name = "NES"
    ) +
    labs(
      x = "Time point", y = NULL,
      title = paste0(outcome_name, ": later vs early KO GSEA"),
      subtitle = subtitle
    ) +
    theme_bw(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.text.y = element_text(size = 8),
      plot.title = element_text(face = "bold")
    )

  list(
    plot = plot,
    height = if (all_significant) {
      max(6, 2 + 0.28 * length(selected_ids))
    } else {
      max(6, min(20, 2 + 0.28 * length(selected_ids)))
    }
  )
}
```

```{r run-outcome-analyses}
outcome_metadata <- metadata_for(colnames(ko_raw), outcome_time_levels)
outcome_ko <- ko_raw[, rownames(outcome_metadata), drop = FALSE]
outcome_ko <- outcome_ko[!grepl("g__", rownames(outcome_ko)), , drop = FALSE]
outcome_ko_cpm <- to_cpm(outcome_ko)
outcome_ko_cpm <- outcome_ko_cpm[
  rowSums(outcome_ko_cpm > 0) > ncol(outcome_ko_cpm) * prevalence_fraction,
  , drop = FALSE
]
outcome_ko_log2 <- log2(outcome_ko_cpm + pseudo_count)

outcome_summary <- list()

for (outcome_name in names(outcome_definitions)) {
  definition <- outcome_definitions[[outcome_name]]
  fitted <- fit_outcome_by_time(
    outcome_ko_log2,
    outcome_metadata,
    outcome_name,
    definition$metadata_column,
    definition$group_map
  )
  enrichment <- run_outcome_gsea(fitted$results)
  enrichment_table <- bind_rows(lapply(names(enrichment), function(timepoint) {
    if (is.null(enrichment[[timepoint]])) return(NULL)
    tab <- as.data.frame(enrichment[[timepoint]])
    tab$timeID <- timepoint
    tab$outcome <- outcome_name
    tab$comparison <- "later_vs_early"
    tab
  }))

  this_dir <- file.path(outcome_dir, outcome_name)
  dir.create(this_dir, recursive = TRUE, showWarnings = FALSE)

  saveRDS(
    fitted$results,
    file.path(this_dir, paste0(outcome_name, ".later_vs_early.differential_KO.rds"))
  )
  write.csv(
    bind_rows(fitted$results),
    file.path(this_dir, paste0(outcome_name, ".later_vs_early.differential_KO.csv")),
    row.names = FALSE
  )
  write.csv(
    fitted$sample_counts,
    file.path(this_dir, paste0(outcome_name, ".later_vs_early.sample_counts.csv")),
    row.names = FALSE
  )
  saveRDS(
    enrichment,
    file.path(this_dir, paste0(outcome_name, ".later_vs_early.KO_GSEA.rds"))
  )
  write.csv(
    enrichment_table,
    file.path(this_dir, paste0(outcome_name, ".later_vs_early.KO_GSEA.csv")),
    row.names = FALSE
  )

  top_plot <- plot_outcome_heatmap(
    enrichment_table,
    outcome_name,
    all_significant = FALSE
  )
  ggsave(
    file.path(this_dir, paste0(outcome_name, ".later_vs_early.KO_GSEA_heatmap.pdf")),
    plot = top_plot$plot,
    width = 10,
    height = top_plot$height,
    limitsize = FALSE
  )

  all_plot <- plot_outcome_heatmap(
    enrichment_table,
    outcome_name,
    all_significant = TRUE
  )
  ggsave(
    file.path(
      this_dir,
      paste0(outcome_name, ".later_vs_early.KO_GSEA_all_significant_heatmap.pdf")
    ),
    plot = all_plot$plot,
    width = 10,
    height = all_plot$height,
    limitsize = FALSE
  )

  outcome_summary[[outcome_name]] <- fitted$sample_counts
}

outcome_summary_table <- bind_rows(outcome_summary)
write.csv(
  outcome_summary_table,
  file.path(table_dir, "outcome_KO_analysis_sample_counts.csv"),
  row.names = FALSE
)
knitr::kable(outcome_summary_table)
```

# Output inventory

```{r output-inventory}
output_inventory <- data.frame(
  file = sort(list.files(output_dir, recursive = TRUE)),
  stringsAsFactors = FALSE
)
write.csv(
  output_inventory,
  file.path(table_dir, "output_inventory.csv"),
  row.names = FALSE
)
knitr::kable(output_inventory)
```
