---
title: "16S Bray-Curtis distances and sample-characteristic distributions"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: true
---

This analysis characterizes Bray-Curtis distances across longitudinal time
points and summarizes the distributions of selected sample and clinical
characteristics. Six figures are generated using predefined group orders,
Wilcoxon tests, color palettes, dimensions, and random seed.

```{r setup, include=FALSE}
options(stringsAsFactors = FALSE)

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(scales)
  library(vegan)
})

input_file <- "000.input16S.Rds"
analysis_name <- "04.16s_BrayCurtis_and_metadata_distributions"
output_root <- analysis_name
figure_dir <- file.path(output_root, "figures")
table_dir <- file.path(output_root, "tables")

dir.create(output_root, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = file.path(output_root, "knitr_figures", "")
)

# Parameters retained from the source analysis.
set.seed(20260529)
timepoint_order <- c("BL", "D0", "D14", "M1", "M2", "M3", "M4", "M5", "M6")
all_distance_groups <- c("BL", "D0", "D14", "M1", "M2", "M3", "M6")
broad_abx_distance_groups <- c("D0", "D14", "M1", "M2")
```

# 1. Reusable functions

```{r functions}
write_table <- function(data, filename) {
  utils::write.csv(
    data,
    file.path(table_dir, filename),
    row.names = FALSE,
    na = ""
  )
}

make_count_table <- function(metadata, variable, category_levels) {
  metadata |>
    dplyr::transmute(
      group_label = .data$group_label,
      category = factor(.data[[variable]], levels = category_levels)
    ) |>
    dplyr::filter(!is.na(.data$group_label), !is.na(.data$category)) |>
    dplyr::count(.data$group_label, .data$category, name = "n") |>
    dplyr::group_by(.data$group_label) |>
    dplyr::mutate(proportion = .data$n / sum(.data$n)) |>
    dplyr::ungroup()
}

plot_count_distribution <- function(count_table, palette, title, legend_title) {
  ggplot2::ggplot(
    count_table,
    ggplot2::aes(x = .data$group_label, y = .data$proportion, fill = .data$category)
  ) +
    ggplot2::geom_col(width = 0.72, colour = "white", linewidth = 0.3) +
    ggplot2::geom_text(
      ggplot2::aes(label = ifelse(.data$n > 0, .data$n, "")),
      position = ggplot2::position_stack(vjust = 0.5),
      size = 3.1,
      colour = "white",
      fontface = "bold",
      family = "ArialMT"
    ) +
    ggplot2::scale_fill_manual(values = palette, drop = FALSE) +
    ggplot2::scale_y_continuous(
      labels = scales::percent_format(accuracy = 1),
      expand = ggplot2::expansion(mult = c(0, 0.04))
    ) +
    ggplot2::labs(
      x = NULL,
      y = "Proportion of samples",
      title = title,
      fill = legend_title
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      text = ggplot2::element_text(family = "ArialMT"),
      panel.grid = ggplot2::element_blank(),
      plot.title = ggplot2::element_text(face = "bold", hjust = 0.5, size = 12),
      legend.position = "top"
    )
}

extract_within_group_distances <- function(
    metadata,
    distance_matrix,
    groups,
    stratification_variable = NULL) {
  output <- list()
  output_index <- 1L

  for (group_name in groups) {
    group_metadata <- metadata |>
      dplyr::filter(.data$group_label == group_name)

    stratum_values <- if (is.null(stratification_variable)) {
      NA_character_
    } else {
      unique(as.character(group_metadata[[stratification_variable]]))
    }

    for (stratum_value in stratum_values) {
      subgroup_metadata <- if (is.null(stratification_variable)) {
        group_metadata
      } else {
        group_metadata |>
          dplyr::filter(.data[[stratification_variable]] == stratum_value)
      }

      sample_ids <- subgroup_metadata$sample_id
      if (length(sample_ids) < 2L) {
        next
      }

      subgroup_matrix <- distance_matrix[
        sample_ids,
        sample_ids,
        drop = FALSE
      ]
      lower_triangle <- lower.tri(subgroup_matrix)

      output[[output_index]] <- data.frame(
        group_label = group_name,
        stratum = stratum_value,
        BC_distance = subgroup_matrix[lower_triangle],
        stringsAsFactors = FALSE
      )
      output_index <- output_index + 1L
    }
  }

  dplyr::bind_rows(output)
}

safe_wilcox_p <- function(x, y) {
  if (length(x) == 0L || length(y) == 0L) {
    return(NA_real_)
  }
  suppressWarnings(stats::wilcox.test(x, y)$p.value)
}

format_plot_p <- function(p) {
  if (is.na(p)) {
    return("NA")
  }
  if (p < 1e-4) {
    return(format(signif(p, 2), scientific = TRUE, trim = TRUE))
  }
  format(signif(p, 2), scientific = FALSE, trim = TRUE)
}

significance_symbol <- function(p) {
  dplyr::case_when(
    is.na(p) ~ "NA",
    p <= 1e-4 ~ "****",
    p <= 1e-3 ~ "***",
    p <= 1e-2 ~ "**",
    p <= 5e-2 ~ "*",
    TRUE ~ "ns"
  )
}
```

# 2. Input preparation and Bray-Curtis distance

```{r prepare-input}
if (!file.exists(input_file)) {
  stop("Input file not found: ", input_file)
}

input16S <- readRDS(input_file)
if (!is.list(input16S) || !all(c("dat_rela", "metainfo") %in% names(input16S))) {
  stop("000.input16S.Rds must contain dat_rela and metainfo.")
}

dat_rela <- input16S$dat_rela
metainfo <- input16S$metainfo

required_metadata <- c(
  "Sample_Name", "timeIDcontiue", "batch", "Disease",
  "abx_recent", "broad_abx_14d"
)
missing_metadata <- setdiff(required_metadata, names(metainfo))
if (length(missing_metadata) > 0L) {
  stop("Missing metadata column(s): ", paste(missing_metadata, collapse = ", "))
}
if (!identical(colnames(dat_rela), as.character(metainfo$Sample_Name))) {
  stop("dat_rela columns are not aligned with metainfo$Sample_Name.")
}

metadata_select <- metainfo |>
  dplyr::transmute(
    sample_id = as.character(.data$Sample_Name),
    group_label = factor(
      sub("^[0-9]+_", "", .data$timeIDcontiue),
      levels = timepoint_order
    ),
    batch = factor(.data$batch, levels = c("batch1", "batch2")),
    Disease = factor(.data$Disease, levels = c("ANCA", "IIM", "SjD", "SLE", "SSc")),
    abx_recent = factor(.data$abx_recent, levels = c("No", "Yes")),
    broad_abx_14d = factor(.data$broad_abx_14d, levels = c("No", "Yes"))
  ) |>
  dplyr::filter(!is.na(.data$group_label)) |>
  droplevels()

dat_rela_select <- dat_rela[, metadata_select$sample_id, drop = FALSE]
dat_rela_select <- dat_rela_select[rowSums(dat_rela_select) > 0, , drop = FALSE]

abundance_matrix <- t(dat_rela_select)
bray_curtis_distance <- vegan::vegdist(abundance_matrix, method = "bray")
distance_matrix <- as.matrix(bray_curtis_distance)

data.frame(
  selected_samples = nrow(metadata_select),
  retained_taxa = nrow(dat_rela_select),
  timepoints = nlevels(metadata_select$group_label)
)
```

# 3. Sample-characteristic distributions

```{r distribution-data}
recent_palette <- c("No" = "#969696", "Yes" = "#c45c1a")
broad_palette <- c("No" = "#969696", "Yes" = "#c45c1a")
batch_palette <- c("batch1" = "#1F77B4", "batch2" = "tomato3")
disease_palette <- c(
  "ANCA" = "grey",
  "IIM" = "#2CA02C",
  "SjD" = "#1F77B4",
  "SLE" = "#D62728",
  "SSc" = "#9467BD"
)

recent_count_table <- make_count_table(
  metadata_select,
  "abx_recent",
  c("No", "Yes")
)
broad_count_table <- make_count_table(
  metadata_select,
  "broad_abx_14d",
  c("No", "Yes")
)
batch_count_table <- make_count_table(
  metadata_select,
  "batch",
  c("batch1", "batch2")
)
disease_count_table <- make_count_table(
  metadata_select,
  "Disease",
  c("ANCA", "IIM", "SjD", "SLE", "SSc")
)

write_table(recent_count_table, "recent_antibiotic_exposure_counts.csv")
write_table(broad_count_table, "recent_broad_spectrum_antibiotic_exposure_counts.csv")
write_table(batch_count_table, "sequencing_batch_counts.csv")
write_table(disease_count_table, "disease_category_counts.csv")
```

## 3.1 Recent antibiotic exposure

```{r recent-antibiotic-plot, fig.width=7.02, fig.height=3.39}
p_recent_antibiotic <- plot_count_distribution(
  recent_count_table,
  recent_palette,
  "Recent antibiotic exposure by time window",
  "Recent antibiotic exposure"
)

ggplot2::ggsave(
  file.path(figure_dir, "recent_antibiotic_exposure_by_timepoint.pdf"),
  p_recent_antibiotic,
  width = 7.02,
  height = 3.39
)

p_recent_antibiotic
```

## 3.2 Recent broad-spectrum antibiotic exposure

```{r recent-broad-antibiotic-plot, fig.width=7.02, fig.height=3.39}
p_recent_broad_antibiotic <- plot_count_distribution(
  broad_count_table,
  broad_palette,
  "Broad-spectrum antibiotic exposure (14d) by time window",
  "Broad-spectrum antibiotic exposure (14d)"
)

ggplot2::ggsave(
  file.path(figure_dir, "recent_broad_spectrum_antibiotic_exposure_by_timepoint.pdf"),
  p_recent_broad_antibiotic,
  width = 7.02,
  height = 3.39
)

p_recent_broad_antibiotic
```

## 3.3 Sequencing batch

```{r sequencing-batch-plot, fig.width=7.02, fig.height=3.39}
p_sequencing_batch <- plot_count_distribution(
  batch_count_table,
  batch_palette,
  "Sequencing batch by time window",
  "Sequencing batch"
)

ggplot2::ggsave(
  file.path(figure_dir, "sequencing_batch_distribution_by_timepoint.pdf"),
  p_sequencing_batch,
  width = 7.02,
  height = 3.39
)

p_sequencing_batch
```

## 3.4 Disease category

```{r disease-category-plot, fig.width=7.02, fig.height=3.39}
p_disease <- plot_count_distribution(
  disease_count_table,
  disease_palette,
  "Disease category by time window",
  "Disease"
)

ggplot2::ggsave(
  file.path(figure_dir, "disease_distribution_by_timepoint.pdf"),
  p_disease,
  width = 7.02,
  height = 3.39
)

p_disease
```

# 4. Bray-Curtis distance by broad-spectrum antibiotic exposure

Pairwise distances are calculated within each timepoint and antibiotic-exposure
stratum. The Wilcoxon test and original grey/tomato palette are retained.

```{r broad-antibiotic-distance-data}
broad_abx_distance_data <- extract_within_group_distances(
  metadata_select,
  distance_matrix,
  broad_abx_distance_groups,
  stratification_variable = "broad_abx_14d"
) |>
  dplyr::mutate(
    group_label = factor(.data$group_label, levels = broad_abx_distance_groups),
    broad_abx_14d = factor(.data$stratum, levels = c("No", "Yes"))
  )

broad_abx_tests <- lapply(broad_abx_distance_groups, function(group_name) {
  group_data <- broad_abx_distance_data |>
    dplyr::filter(.data$group_label == group_name)
  no_distances <- group_data$BC_distance[group_data$broad_abx_14d == "No"]
  yes_distances <- group_data$BC_distance[group_data$broad_abx_14d == "Yes"]
  p_value <- safe_wilcox_p(no_distances, yes_distances)

  data.frame(
    group_label = group_name,
    no_pairwise_distances = length(no_distances),
    yes_pairwise_distances = length(yes_distances),
    median_no = if (length(no_distances) > 0L) stats::median(no_distances) else NA_real_,
    median_yes = if (length(yes_distances) > 0L) stats::median(yes_distances) else NA_real_,
    P_value = p_value,
    significance = significance_symbol(p_value),
    stringsAsFactors = FALSE
  )
}) |>
  dplyr::bind_rows() |>
  dplyr::mutate(
    group_label = factor(.data$group_label, levels = broad_abx_distance_groups),
    label_y = vapply(
      as.character(.data$group_label),
      function(group_name) {
        max(
          broad_abx_distance_data$BC_distance[
            broad_abx_distance_data$group_label == group_name
          ],
          na.rm = TRUE
        ) * 1.03
      },
      numeric(1)
    )
  )

write_table(
  broad_abx_tests |>
    dplyr::select(-dplyr::all_of("label_y")),
  "bray_curtis_broad_spectrum_antibiotic_tests.csv"
)
```

```{r broad-antibiotic-distance-plot, fig.width=5.30, fig.height=3.39}
p_broad_abx_distance <- ggplot2::ggplot(
  broad_abx_distance_data,
  ggplot2::aes(
    x = .data$group_label,
    y = .data$BC_distance,
    fill = .data$broad_abx_14d
  )
) +
  ggplot2::geom_boxplot(
    width = 0.6,
    color = "grey",
    alpha = 0.85,
    linewidth = 0.6,
    position = ggplot2::position_dodge(0.75)
  ) +
  ggplot2::geom_text(
    data = broad_abx_tests,
    ggplot2::aes(
      x = .data$group_label,
      y = .data$label_y,
      label = .data$significance
    ),
    inherit.aes = FALSE,
    vjust = 0,
    size = 3.5,
    family = "ArialMT"
  ) +
  ggplot2::scale_fill_manual(values = c("No" = "#969696", "Yes" = "tomato3")) +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0.02, 0.12))) +
  ggplot2::labs(
    y = "Pairwise within-group Bray-Curtis distance",
    x = NULL,
    fill = "14d Broad Abx"
  ) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    aspect.ratio = 1,
    legend.position = "none",
    legend.title = ggplot2::element_blank(),
    panel.grid = ggplot2::element_blank(),
    panel.border = ggplot2::element_rect(fill = NA, colour = "grey87")
  )

ggplot2::ggsave(
  file.path(figure_dir, "bray_curtis_distance_by_broad_spectrum_antibiotic.pdf"),
  p_broad_abx_distance,
  width = 5.30,
  height = 3.39
)

p_broad_abx_distance
```

# 5. Bray-Curtis distance across the full analysis cycle

The source code generated every within-timepoint distance twice because it
looped over antibiotic status without filtering on that status. This redundant
duplication is removed here. The duplicated BL-D14 comparison in the source
annotation list is also retained only once.

```{r all-timepoint-distance-data}
all_timepoint_distance_data <- extract_within_group_distances(
  metadata_select,
  distance_matrix,
  all_distance_groups
) |>
  dplyr::mutate(
    group_label = factor(.data$group_label, levels = all_distance_groups)
  )

all_distance_summary <- all_timepoint_distance_data |>
  dplyr::group_by(.data$group_label) |>
  dplyr::summarise(
    n_pairwise_distances = dplyr::n(),
    median_BC_distance = stats::median(.data$BC_distance),
    Q1_BC_distance = stats::quantile(.data$BC_distance, 0.25),
    Q3_BC_distance = stats::quantile(.data$BC_distance, 0.75),
    .groups = "drop"
  )

comparison_definitions <- data.frame(
  group_1 = c("BL", "D0", "BL", "D14", "M1", "M2", "BL", "BL", "BL"),
  group_2 = c("D0", "D14", "D14", "M1", "M2", "M3", "M1", "M2", "M3"),
  height_rank = seq_len(9L),
  stringsAsFactors = FALSE
)

distance_range <- diff(range(all_timepoint_distance_data$BC_distance, na.rm = TRUE))
base_annotation_y <- max(all_timepoint_distance_data$BC_distance, na.rm = TRUE) +
  0.035 * distance_range
annotation_step <- 0.09 * distance_range

all_timepoint_tests <- lapply(seq_len(nrow(comparison_definitions)), function(index) {
  group_1 <- comparison_definitions$group_1[index]
  group_2 <- comparison_definitions$group_2[index]
  x <- all_timepoint_distance_data$BC_distance[
    all_timepoint_distance_data$group_label == group_1
  ]
  y <- all_timepoint_distance_data$BC_distance[
    all_timepoint_distance_data$group_label == group_2
  ]
  p_value <- safe_wilcox_p(x, y)

  data.frame(
    group_1 = group_1,
    group_2 = group_2,
    n_pairwise_distances_group_1 = length(x),
    n_pairwise_distances_group_2 = length(y),
    P_value = p_value,
    P_label = format_plot_p(p_value),
    x_start = match(group_1, all_distance_groups),
    x_end = match(group_2, all_distance_groups),
    annotation_y = base_annotation_y + (index - 1L) * annotation_step,
    stringsAsFactors = FALSE
  )
}) |>
  dplyr::bind_rows()

write_table(all_distance_summary, "bray_curtis_all_timepoints_summary.csv")
write_table(
  all_timepoint_tests |>
    dplyr::select(
      .data$group_1,
      .data$group_2,
      .data$n_pairwise_distances_group_1,
      .data$n_pairwise_distances_group_2,
      .data$P_value
    ),
  "bray_curtis_all_timepoints_wilcoxon_tests.csv"
)
```

```{r all-timepoint-distance-plot, fig.width=5.38, fig.height=4.85}
bracket_tip <- 0.018 * distance_range

p_all_timepoint_distance <- ggplot2::ggplot(
  all_timepoint_distance_data,
  ggplot2::aes(x = .data$group_label, y = .data$BC_distance)
) +
  ggplot2::geom_boxplot(
    width = 0.6,
    color = "grey",
    alpha = 0.85,
    linewidth = 0.6
  ) +
  ggplot2::geom_segment(
    data = all_timepoint_tests,
    ggplot2::aes(
      x = .data$x_start,
      xend = .data$x_end,
      y = .data$annotation_y,
      yend = .data$annotation_y
    ),
    inherit.aes = FALSE,
    linewidth = 0.35
  ) +
  ggplot2::geom_segment(
    data = all_timepoint_tests,
    ggplot2::aes(
      x = .data$x_start,
      xend = .data$x_start,
      y = .data$annotation_y,
      yend = .data$annotation_y - bracket_tip
    ),
    inherit.aes = FALSE,
    linewidth = 0.35
  ) +
  ggplot2::geom_segment(
    data = all_timepoint_tests,
    ggplot2::aes(
      x = .data$x_end,
      xend = .data$x_end,
      y = .data$annotation_y,
      yend = .data$annotation_y - bracket_tip
    ),
    inherit.aes = FALSE,
    linewidth = 0.35
  ) +
  ggplot2::geom_text(
    data = all_timepoint_tests,
    ggplot2::aes(
      x = (.data$x_start + .data$x_end) / 2,
      y = .data$annotation_y,
      label = .data$P_label
    ),
    inherit.aes = FALSE,
    vjust = -0.25,
    size = 3.2,
    family = "ArialMT"
  ) +
  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(mult = c(0.02, 0.03))
  ) +
  ggplot2::labs(
    y = "Pairwise within-group Bray-Curtis distance",
    x = NULL
  ) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    aspect.ratio = 1,
    legend.position = "none",
    panel.grid = ggplot2::element_blank(),
    panel.border = ggplot2::element_rect(fill = NA, colour = "grey87")
  )

ggplot2::ggsave(
  file.path(figure_dir, "bray_curtis_distance_all_timepoints.pdf"),
  p_all_timepoint_distance,
  width = 5.38,
  height = 4.85
)

p_all_timepoint_distance
```

# 6. Session information

```{r session-information}
sessionInfo()
```
