---
title: "LEfSe comparisons with baseline across treatment timepoints"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: true
---

This analysis identifies bacterial species that differ between baseline and
six post-treatment timepoints using LEfSe, visualizes their abundance patterns,
summarizes recurrent genera, and compares two Bifidobacterium species over time.

```{r setup, include=FALSE}
options(stringsAsFactors = FALSE)

suppressPackageStartupMessages({
  library(lefser)
  library(SummarizedExperiment)
  library(ComplexHeatmap)
  library(circlize)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(forcats)
  library(ggplot2)
  library(rstatix)
  library(ggpubr)
})

input_file <- "000.input16S.Rds"
taxonomy_file <- "000.lefse_taxonomy_annotation.Rds"
analysis_name <- "05.lefseCompareWithBaseline"
output_root <- analysis_name
figure_dir <- file.path(output_root, "figures")
table_dir <- file.path(output_root, "tables")

dir.create(output_root, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = file.path(output_root, "knitr_figures", "")
)

baseline_timepoint <- "001_BL"
comparison_timepoints <- c(
  "002_D0", "003_D14", "004_M1",
  "005_M2", "006_M3", "009_M6"
)
analysis_timepoints <- c(
  "001_BL", "002_D0", "003_D14", "004_M1", "005_M2",
  "006_M3", "007_M4", "008_M5", "009_M6"
)

timepoint_colors <- c(
  "001_BL" = "#999999",
  "002_D0" = "#377EB8",
  "003_D14" = "#E41A1C",
  "004_M1" = "#984EA3",
  "005_M2" = "#FF7F00",
  "006_M3" = "#FFFF33",
  "007_M4" = "#A65628",
  "008_M5" = "#F781BF",
  "009_M6" = "#4DAF4A"
)
```

# 1. Input preparation

```{r prepare-input}
if (!file.exists(input_file)) {
  stop("Input file not found: ", input_file)
}
if (!file.exists(taxonomy_file)) {
  stop("Taxonomy annotation file not found: ", taxonomy_file)
}

input16S <- readRDS(input_file)
if (!is.list(input16S) || !all(c("dat_rela", "metainfo") %in% names(input16S))) {
  stop("000.input16S.Rds must contain dat_rela and metainfo.")
}

dat_rela <- input16S$dat_rela
metainfo <- input16S$metainfo
dat_name <- readRDS(taxonomy_file)

required_metadata <- c("Sample_Name", "timeIDcontiue")
missing_metadata <- setdiff(required_metadata, names(metainfo))
if (length(missing_metadata) > 0L) {
  stop("Missing metadata column(s): ", paste(missing_metadata, collapse = ", "))
}
if (!identical(colnames(dat_rela), as.character(metainfo$Sample_Name))) {
  stop("dat_rela columns are not aligned with metainfo$Sample_Name.")
}
if (!"species" %in% names(dat_name)) {
  stop("The taxonomy annotation must contain a species column.")
}

metadata_select <- metainfo |>
  dplyr::filter(.data$timeIDcontiue %in% analysis_timepoints) |>
  dplyr::mutate(
    Sample_Name = as.character(.data$Sample_Name),
    timeIDcontiue = as.character(.data$timeIDcontiue)
  )

dat_rela_select <- dat_rela[, metadata_select$Sample_Name, drop = FALSE]
dat_rela_select <- dat_rela_select[rowSums(dat_rela_select) > 0, , drop = FALSE]

dat_name <- dat_name[match(rownames(dat_rela_select), dat_name$species), , drop = FALSE]
if (anyNA(dat_name$species)) {
  stop("Some abundance rows are missing from the taxonomy annotation.")
}
rownames(dat_name) <- dat_name$species
if (!identical(rownames(dat_name), rownames(dat_rela_select))) {
  stop("Taxonomy rows are not aligned with the abundance matrix.")
}

rownames(metadata_select) <- metadata_select$Sample_Name
taxonomy_row_data <- dat_name[, setdiff(names(dat_name), "merger"), drop = FALSE]

objectD <- SummarizedExperiment::SummarizedExperiment(
  assays = list(counts = as.matrix(dat_rela_select)),
  colData = as.data.frame(metadata_select),
  rowData = taxonomy_row_data
)

data.frame(
  samples = ncol(objectD),
  species = nrow(objectD),
  timepoints = dplyr::n_distinct(metadata_select$timeIDcontiue)
)
```

# 2. LEfSe comparisons with baseline

Each post-treatment timepoint is compared independently with baseline. The
original random seed is reset before every LEfSe fit.

```{r run-lefse}
lefse_results <- list()

for (timepoint in comparison_timepoints) {
  object_select <- objectD[
    , objectD$timeIDcontiue %in% c(baseline_timepoint, timepoint)
  ]

  set.seed(1234)
  result_timepoint <- suppressWarnings(
    lefser::lefser(
      object_select,
      groupCol = "timeIDcontiue"
    )
  )

  result_timepoint <- as.data.frame(result_timepoint)
  if (nrow(result_timepoint) > 0L) {
    result_timepoint$group <- ifelse(
      result_timepoint$scores > 0,
      "TimePoint",
      "BL"
    )
  }
  lefse_results[[timepoint]] <- result_timepoint
}

lefse_result_summary <- data.frame(
  timepoint = comparison_timepoints,
  significant_species = vapply(
    comparison_timepoints,
    function(timepoint) nrow(lefse_results[[timepoint]]),
    integer(1)
  )
)

lefse_result_summary
```

# 3. LEfSe score and abundance heatmap

```{r prepare-heatmap-data}
lefse_bacteria <- unique(unlist(lapply(
  comparison_timepoints,
  function(timepoint) as.character(lefse_results[[timepoint]]$Names)
)))

counts_matrix <- SummarizedExperiment::assay(objectD, "counts")
metadata <- as.data.frame(SummarizedExperiment::colData(objectD))

target_bacteria <- intersect(lefse_bacteria, rownames(counts_matrix))
valid_samples <- intersect(metadata$Sample_Name, colnames(counts_matrix))
heatmap_data <- counts_matrix[target_bacteria, valid_samples, drop = FALSE]

matched_index <- match(colnames(heatmap_data), metadata$Sample_Name)
metadata_matched <- metadata[matched_index, , drop = FALSE]

log_cpm <- log2(heatmap_data + 1)
calculate_z_score <- function(values) {
  value_sd <- stats::sd(values)
  if (!is.finite(value_sd) || value_sd == 0) {
    return(rep(0, length(values)))
  }
  (values - mean(values)) / value_sd
}
heatmap_data_z <- t(apply(log_cpm, 1, calculate_z_score))
heatmap_data_z[heatmap_data_z > 3] <- 3
heatmap_data_z[heatmap_data_z < -3] <- -3

lefse_score_matrix <- matrix(
  NA_real_,
  nrow = length(target_bacteria),
  ncol = length(comparison_timepoints),
  dimnames = list(target_bacteria, comparison_timepoints)
)

for (timepoint in comparison_timepoints) {
  result_timepoint <- lefse_results[[timepoint]]
  match_index <- match(target_bacteria, as.character(result_timepoint$Names))
  lefse_score_matrix[, timepoint] <- result_timepoint$scores[match_index]
}

lefse_score_output <- data.frame(
  Names = rownames(lefse_score_matrix),
  lefse_score_matrix,
  check.names = FALSE
)
utils::write.csv(
  lefse_score_output,
  file.path(table_dir, "lefse_score_matrix.csv"),
  row.names = FALSE,
  na = ""
)
```

```{r draw-heatmap, fig.width=9.95, fig.height=9.17}
minimum_score <- min(lefse_score_matrix, na.rm = TRUE)
maximum_score <- max(lefse_score_matrix, na.rm = TRUE)

lda_color_function <- if (minimum_score < 0) {
  circlize::colorRamp2(
    c(minimum_score, 0, maximum_score),
    c("seagreen", "white", "purple")
  )
} else {
  circlize::colorRamp2(c(0, maximum_score), c("white", "purple"))
}

lefse_score_heatmap <- ComplexHeatmap::Heatmap(
  lefse_score_matrix,
  name = "LDA Score",
  col = lda_color_function,
  na_col = "grey90",
  rect_gp = grid::gpar(col = "white", lwd = 1),
  cluster_columns = FALSE,
  cluster_rows = TRUE,
  show_row_names = TRUE,
  row_names_gp = grid::gpar(fontsize = 8),
  column_names_rot = 45,
  column_names_gp = grid::gpar(fontsize = 9),
  column_title = "LEfSe Scores",
  width = grid::unit(4, "cm")
)

time_group <- as.character(metadata_matched$timeIDcontiue)
column_annotation <- ComplexHeatmap::HeatmapAnnotation(
  TimePoint = time_group,
  col = list(TimePoint = timepoint_colors)
)
zscore_color_function <- circlize::colorRamp2(
  c(-3, 0, 3),
  c("#084594", "white", "firebrick3")
)

abundance_heatmap <- ComplexHeatmap::Heatmap(
  heatmap_data_z,
  name = "Z-score",
  col = zscore_color_function,
  top_annotation = column_annotation,
  column_split = factor(time_group, levels = names(timepoint_colors)),
  clustering_method_rows = "ward.D",
  clustering_method_columns = "mcquitty",
  cluster_column_slices = FALSE,
  cluster_columns = TRUE,
  show_column_names = FALSE,
  show_row_names = FALSE,
  column_title = "Sample Expressions"
)

combined_heatmap <- abundance_heatmap + lefse_score_heatmap
heatmap_pdf <- file.path(figure_dir, "lefse_score_and_abundance_heatmap.pdf")

grDevices::pdf(heatmap_pdf, width = 9.95, height = 9.17)
ComplexHeatmap::draw(combined_heatmap, main_heatmap = "Z-score")
grDevices::dev.off()

ComplexHeatmap::draw(combined_heatmap, main_heatmap = "Z-score")
```

# 4. Top genera before and after deduplication

```{r prepare-genus-counts}
names_raw <- unlist(lapply(
  comparison_timepoints,
  function(timepoint) as.character(lefse_results[[timepoint]]$Names)
))
names_unique <- unique(names_raw)

process_taxa_count <- function(taxa_names, status_name) {
  genus_extracted <- vapply(
    strsplit(taxa_names, "_", fixed = TRUE),
    function(parts) {
      if (length(parts) >= 3L) parts[3] else NA_character_
    },
    character(1)
  )
  genus_extracted[
    is.na(genus_extracted) |
      genus_extracted == "" |
      genus_extracted == "NA"
  ] <- "Unclassified"

  data.frame(Genus = genus_extracted, stringsAsFactors = FALSE) |>
    dplyr::count(.data$Genus, name = "Count") |>
    dplyr::mutate(Status = status_name)
}

genus_count_raw <- process_taxa_count(names_raw, "Before deduplication")
genus_count_unique <- process_taxa_count(names_unique, "After deduplication")

top10_genera <- genus_count_raw |>
  dplyr::filter(.data$Genus != "Unclassified") |>
  dplyr::slice_max(order_by = .data$Count, n = 10, with_ties = FALSE) |>
  dplyr::pull(.data$Genus)

genus_plot_data <- dplyr::bind_rows(genus_count_raw, genus_count_unique) |>
  dplyr::filter(.data$Genus %in% top10_genera) |>
  dplyr::mutate(
    Status = factor(
      .data$Status,
      levels = c("Before deduplication", "After deduplication")
    )
  ) |>
  tidyr::complete(
    Genus = top10_genera,
    Status = factor(
      c("Before deduplication", "After deduplication"),
      levels = c("Before deduplication", "After deduplication")
    ),
    fill = list(Count = 0)
  )

genus_order <- genus_plot_data |>
  dplyr::filter(.data$Status == "Before deduplication") |>
  dplyr::arrange(.data$Count) |>
  dplyr::pull(.data$Genus)

genus_plot_data <- genus_plot_data |>
  dplyr::mutate(Genus = factor(.data$Genus, levels = genus_order))

genus_line_data <- genus_plot_data |>
  dplyr::select(.data$Genus, .data$Status, .data$Count) |>
  tidyr::pivot_wider(
    names_from = .data$Status,
    values_from = .data$Count,
    values_fill = 0
  )
```

```{r draw-genus-counts, fig.width=5.54, fig.height=5.07}
p_genus_counts <- ggplot2::ggplot() +
  ggplot2::geom_segment(
    data = genus_line_data,
    ggplot2::aes(
      x = .data[["After deduplication"]],
      xend = .data[["Before deduplication"]],
      y = .data$Genus,
      yend = .data$Genus
    ),
    linewidth = 0.7,
    color = "grey70"
  ) +
  ggplot2::geom_point(
    data = genus_plot_data,
    ggplot2::aes(
      x = .data$Count,
      y = .data$Genus,
      color = .data$Status,
      shape = .data$Status
    ),
    size = 4,
    stroke = 1
  ) +
  ggplot2::geom_text(
    data = genus_plot_data,
    ggplot2::aes(
      x = .data$Count + 1,
      y = .data$Genus,
      label = .data$Count,
      color = .data$Status
    ),
    hjust = -0.4,
    size = 3.5,
    show.legend = FALSE
  ) +
  ggplot2::scale_shape_manual(
    values = c("Before deduplication" = 16, "After deduplication" = 17)
  ) +
  ggplot2::scale_color_manual(
    values = c(
      "Before deduplication" = "#D55E00",
      "After deduplication" = "#0072B2"
    )
  ) +
  ggplot2::scale_x_continuous(
    expand = ggplot2::expansion(mult = c(0.02, 0.15)),
    limits = c(0, 10)
  ) +
  ggplot2::labs(
    x = "Count",
    y = "Top 10 genera",
    color = NULL,
    shape = NULL
  ) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    legend.title = ggplot2::element_blank(),
    plot.title = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_text(angle = 0, hjust = 0.5, family = "ArialMT"),
    axis.text = ggplot2::element_text(family = "ArialMT"),
    axis.title = ggplot2::element_text(family = "ArialMT"),
    panel.grid = ggplot2::element_blank(),
    panel.border = ggplot2::element_rect(fill = NA, colour = "grey87")
  )

ggplot2::ggsave(
  file.path(figure_dir, "top10_genera_deduplicated_comparison.pdf"),
  p_genus_counts,
  width = 5.54,
  height = 5.07
)

p_genus_counts
```

# 5. Longitudinal Bifidobacterium comparison

```{r prepare-bifidobacterium-data}
bifidobacterium_species <- c(
  "s__Bifidobacterium_breve",
  "s__Bifidobacterium_dentium"
)
bifidobacterium_timepoints <- c(
  "001_BL", "002_D0", "003_D14",
  "004_M1", "005_M2", "006_M3", "009_M6"
)

missing_species <- setdiff(bifidobacterium_species, rownames(counts_matrix))
if (length(missing_species) > 0L) {
  stop("Missing target species: ", paste(missing_species, collapse = ", "))
}

bifidobacterium_expression <- t(
  counts_matrix[bifidobacterium_species, , drop = FALSE]
) |>
  as.data.frame() |>
  tibble::rownames_to_column("Sample_Name")

bifidobacterium_plot_data <- bifidobacterium_expression |>
  dplyr::inner_join(
    metadata |>
      dplyr::select(.data$Sample_Name, .data$timeIDcontiue),
    by = "Sample_Name"
  ) |>
  dplyr::filter(.data$timeIDcontiue %in% bifidobacterium_timepoints) |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(bifidobacterium_species),
    names_to = "Species",
    values_to = "Abundance"
  ) |>
  dplyr::mutate(
    Log_Abundance = log(as.numeric(.data$Abundance) + 1),
    timeIDcontiue = factor(
      as.character(.data$timeIDcontiue),
      levels = bifidobacterium_timepoints,
      labels = c("BL", "D0", "D14", "M1", "M2", "M3", "M6")
    ),
    Species = factor(.data$Species, levels = bifidobacterium_species)
  )

bifidobacterium_annotation_caps <- bifidobacterium_plot_data |>
  dplyr::group_by(.data$timeIDcontiue) |>
  dplyr::summarise(
    annotation_cap = max(.data$Log_Abundance, na.rm = TRUE) + 0.6,
    .groups = "drop"
  )

bifidobacterium_stat_test <- bifidobacterium_plot_data |>
  dplyr::group_by(.data$timeIDcontiue) |>
  rstatix::wilcox_test(Log_Abundance ~ Species, paired = TRUE) |>
  rstatix::adjust_pvalue(method = "BH") |>
  rstatix::add_significance("p") |>
  rstatix::add_xy_position(x = "timeIDcontiue", dodge = 0.8) |>
  dplyr::left_join(
    bifidobacterium_annotation_caps,
    by = "timeIDcontiue"
  ) |>
  dplyr::mutate(
    y.position = pmin(.data$y.position, .data$annotation_cap)
  ) |>
  dplyr::select(-dplyr::all_of("annotation_cap"))

bifidobacterium_stat_test
```

```{r draw-bifidobacterium-comparison, fig.width=10.14, fig.height=7}
p_bifidobacterium <- ggplot2::ggplot(
  bifidobacterium_plot_data,
  ggplot2::aes(
    x = .data$timeIDcontiue,
    y = .data$Log_Abundance,
    fill = .data$Species
  )
) +
  ggplot2::geom_boxplot(
    position = ggplot2::position_dodge(width = 0.8),
    width = 0.65,
    alpha = 0.65,
    outlier.shape = NA,
    color = "black"
  ) +
  ggplot2::geom_point(
    ggplot2::aes(color = .data$Species),
    position = ggplot2::position_jitterdodge(
      jitter.width = 0.12,
      dodge.width = 0.8,
      seed = 1234
    ),
    size = 1.8,
    alpha = 0.65
  ) +
  ggplot2::stat_summary(
    ggplot2::aes(group = .data$Species, color = .data$Species),
    fun = mean,
    geom = "line",
    linewidth = 1,
    position = ggplot2::position_dodge(width = 0.8)
  ) +
  ggplot2::stat_summary(
    ggplot2::aes(group = .data$Species, color = .data$Species),
    fun = mean,
    geom = "point",
    shape = 21,
    size = 3,
    stroke = 1,
    position = ggplot2::position_dodge(width = 0.8)
  ) +
  ggpubr::stat_compare_means(
    method = "t.test",
    label = "p",
    bracket.size = 0.4,
    label.y = max(bifidobacterium_plot_data$Log_Abundance, na.rm = TRUE) * 1.18
  ) +
  ggpubr::stat_pvalue_manual(
    bifidobacterium_stat_test,
    label = "p.signif",
    tip.length = 0.01,
    bracket.size = 0.4,
    hide.ns = FALSE,
    inherit.aes = FALSE
  ) +
  ggplot2::labs(
    x = NULL,
    y = "Log abundance",
    fill = "Species",
    color = "Species"
  ) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    legend.title = ggplot2::element_blank(),
    plot.title = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_text(angle = 0, hjust = 0.5, family = "ArialMT"),
    axis.text = ggplot2::element_text(family = "ArialMT"),
    axis.title = ggplot2::element_text(family = "ArialMT"),
    panel.grid = ggplot2::element_blank(),
    panel.border = ggplot2::element_rect(fill = NA, colour = "grey87")
  ) +
  ggplot2::scale_fill_manual(values = c("#ef6548", "#67a9cf")) +
  ggplot2::scale_color_manual(values = c("#ef6548", "#67a9cf")) +
  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(mult = c(0.02, 0.08))
  )

ggplot2::ggsave(
  file.path(figure_dir, "bifidobacterium_species_comparison.pdf"),
  p_bifidobacterium,
  width = 10.14,
  height = 7
)

p_bifidobacterium
```

# 6. Session information

```{r session-information}
sessionInfo()
```
