---
title: "Gut bacterial features associated with favorable B-cell subsets"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: true
---

This analysis evaluates associations between longitudinal gut bacterial
profiles and the combined proportion of naive B cells, immature B cells, and
unswitched memory B cells. Samples are stratified at the median B-cell
proportion for LEfSe and species-level abundance comparisons.

```{r setup, include=FALSE}
options(stringsAsFactors = FALSE)

suppressPackageStartupMessages({
  library(lefser)
  library(SummarizedExperiment)
  library(dplyr)
  library(ggplot2)
  library(ggpubr)
})

input_file <- "000.input16S.Rds"
cell_metric_file <- "000.good_B_cell_metrics.csv"
taxonomy_file <- "000.lefse_taxonomy_annotation.Rds"

analysis_name <- "06.lefse_Bsub_clone"
output_root <- analysis_name
figure_dir <- file.path(output_root, "figures")
table_dir <- file.path(output_root, "tables")

dir.create(output_root, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = file.path(output_root, "knitr_figures", "")
)

excluded_timepoints <- c("001_BL", "002_D0", "003_D14")
excluded_disease <- "IIM"
prevalence_fraction <- 0.10
lefse_seed <- 1234

target_boxplot_species <- c(
  "s__Blautia_wexlerae",
  "s__Blautia_caecimuris",
  "s__Ruminococcus_bromii",
  "s__Romboutsia_ilealis"
)
```

# 1. Input preparation

```{r prepare-input}
required_files <- c(input_file, cell_metric_file, taxonomy_file)
missing_files <- required_files[!file.exists(required_files)]
if (length(missing_files) > 0L) {
  stop("Missing input file(s): ", paste(missing_files, collapse = ", "))
}

input16S <- readRDS(input_file)
if (!is.list(input16S) || !all(c("dat_rela", "metainfo") %in% names(input16S))) {
  stop("000.input16S.Rds must contain dat_rela and metainfo.")
}

dat_rela <- input16S$dat_rela
metainfo <- input16S$metainfo
cell_metrics <- utils::read.csv(cell_metric_file, check.names = FALSE)
dat_name <- readRDS(taxonomy_file)

required_metadata <- c("Sample_Name", "Disease", "timeIDcontiue")
missing_metadata <- setdiff(required_metadata, names(metainfo))
if (length(missing_metadata) > 0L) {
  stop("Missing metadata column(s): ", paste(missing_metadata, collapse = ", "))
}
required_cell_metrics <- c(
  "Sample_Name", "Good_B_cell_count",
  "Total_cell_count", "Good_B_cell_percent"
)
missing_cell_metrics <- setdiff(required_cell_metrics, names(cell_metrics))
if (length(missing_cell_metrics) > 0L) {
  stop("Missing B-cell metric column(s): ", paste(missing_cell_metrics, collapse = ", "))
}
if (!identical(colnames(dat_rela), as.character(metainfo$Sample_Name))) {
  stop("dat_rela columns are not aligned with metainfo$Sample_Name.")
}
if (!"species" %in% names(dat_name)) {
  stop("The taxonomy annotation must contain a species column.")
}

metadata_select <- metainfo |>
  dplyr::left_join(cell_metrics, by = "Sample_Name") |>
  dplyr::filter(
    .data$Disease != excluded_disease,
    !.data$timeIDcontiue %in% excluded_timepoints,
    is.finite(.data$Good_B_cell_percent)
  ) |>
  dplyr::mutate(
    Sample_Name = as.character(.data$Sample_Name),
    B_cell_group = ifelse(
      .data$Good_B_cell_percent > stats::median(.data$Good_B_cell_percent),
      "more",
      "less"
    ),
    B_cell_group = factor(.data$B_cell_group, levels = c("less", "more"))
  )

if (nrow(metadata_select) == 0L) {
  stop("No samples remain after metadata and B-cell filtering.")
}

dat_rela_select <- dat_rela[, metadata_select$Sample_Name, drop = FALSE]
dat_rela_select <- dat_rela_select[rowSums(dat_rela_select) > 0, , drop = FALSE]

dat_name <- dat_name[match(rownames(dat_rela_select), dat_name$species), , drop = FALSE]
if (anyNA(dat_name$species)) {
  stop("Some abundance rows are missing from the taxonomy annotation.")
}
rownames(dat_name) <- dat_name$species
if (!identical(rownames(dat_name), rownames(dat_rela_select))) {
  stop("Taxonomy rows are not aligned with the abundance matrix.")
}

missing_boxplot_species <- setdiff(
  target_boxplot_species,
  rownames(dat_rela_select)
)
if (length(missing_boxplot_species) > 0L) {
  stop("Missing boxplot species: ", paste(missing_boxplot_species, collapse = ", "))
}

rownames(metadata_select) <- metadata_select$Sample_Name

data.frame(
  samples = nrow(metadata_select),
  species = nrow(dat_rela_select),
  median_Good_B_cell_percent = stats::median(metadata_select$Good_B_cell_percent),
  less_group = sum(metadata_select$B_cell_group == "less"),
  more_group = sum(metadata_select$B_cell_group == "more")
)
```

# 2. Species correlations with the favorable B-cell proportion

Species detected in more than 10% of selected samples are retained for the
Spearman correlation analysis. Correlation coefficients are used as labels in
the LEfSe figure when the nominal correlation P value is below 0.05.

```{r species-correlations}
prevalent_species <- apply(
  dat_rela_select,
  1,
  function(values) {
    sum(values > 0) > ncol(dat_rela_select) * prevalence_fraction
  }
)
correlation_matrix <- dat_rela_select[prevalent_species, , drop = FALSE]

spearman_rho <- apply(
  correlation_matrix,
  1,
  function(values) {
    stats::cor(
      values,
      metadata_select$Good_B_cell_percent,
      method = "spearman",
      use = "pairwise.complete.obs"
    )
  }
)
spearman_p <- apply(
  correlation_matrix,
  1,
  function(values) {
    suppressWarnings(
      stats::cor.test(
        values,
        metadata_select$Good_B_cell_percent,
        method = "spearman"
      )$p.value
    )
  }
)

correlation_results <- data.frame(
  Names = rownames(correlation_matrix),
  spearman_rho = as.numeric(spearman_rho),
  spearman_p = as.numeric(spearman_p),
  stringsAsFactors = FALSE
) |>
  dplyr::arrange(.data$spearman_p)

significant_correlations <- correlation_results |>
  dplyr::filter(.data$spearman_p < 0.05) |>
  dplyr::arrange(.data$spearman_rho)

data.frame(
  prevalent_species = nrow(correlation_results),
  nominally_significant_correlations = nrow(significant_correlations)
)
```

# 3. LEfSe comparison of high and low B-cell groups

```{r run-lefse}
taxonomy_row_data <- dat_name[, setdiff(names(dat_name), "merger"), drop = FALSE]

lefse_object <- SummarizedExperiment::SummarizedExperiment(
  assays = list(counts = as.matrix(dat_rela_select)),
  colData = as.data.frame(metadata_select),
  rowData = taxonomy_row_data
)

set.seed(lefse_seed)
lefse_result <- suppressWarnings(
  lefser::lefser(
    lefse_object,
    groupCol = "B_cell_group"
  )
) |>
  as.data.frame()

lefse_result$class <- dat_name$class[
  match(lefse_result$Names, dat_name$species)
]
lefse_result$correlation_label <- significant_correlations$spearman_rho[
  match(lefse_result$Names, significant_correlations$Names)
]

lefse_output <- lefse_result |>
  dplyr::select(.data$Names, .data$scores, .data$class)

utils::write.csv(
  lefse_output,
  file.path(table_dir, "lefse_good_B_cell_groups.csv"),
  row.names = FALSE,
  na = ""
)

lefse_output
```

```{r plot-lefse, fig.width=10.11, fig.height=4.16}
lefse_plot_data <- lefse_result |>
  dplyr::mutate(
    class = factor(
      .data$class,
      levels = c(
        "c__Gammaproteobacteria",
        "c__Bacilli",
        "c__Clostridia"
      )
    )
  )

p_lefse <- ggpubr::ggdotchart(
  lefse_plot_data,
  x = "Names",
  y = "scores",
  color = "class",
  sorting = "ascending",
  group = "class",
  sort.by.groups = TRUE,
  add = "segments",
  palette = c("#00AFBB", "#E7B800", "#FC4E07"),
  label = round(lefse_plot_data$correlation_label, 2),
  dot.size = 6
) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    legend.title = ggplot2::element_blank(),
    axis.title.x = ggplot2::element_blank(),
    plot.title = ggplot2::element_blank(),
    axis.text = ggplot2::element_text(family = "ArialMT", angle = 90, hjust = 1),
    axis.title = ggplot2::element_text(family = "ArialMT"),
    panel.grid = ggplot2::element_blank(),
    panel.border = ggplot2::element_rect(fill = NA, colour = "grey87")
  ) +
  ggplot2::scale_y_continuous(expand = c(0.2, 0.2)) +
  ggplot2::labs(y = "LDA score")

ggplot2::ggsave(
  file.path(figure_dir, "lefse_good_B_cell_groups.pdf"),
  p_lefse,
  width = 10.11,
  height = 4.16
)

p_lefse
```

# 4. Species-level abundance comparisons

```{r boxplot-functions}
group_colors <- c("less" = "#67a9cf", "more" = "#ef8a62")

make_species_boxplot <- function(species_name) {
  plot_data <- data.frame(
    B_cell_group = metadata_select$B_cell_group,
    abundance = as.numeric(dat_rela_select[species_name, ]),
    stringsAsFactors = FALSE
  )

  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(x = .data$B_cell_group, y = .data$abundance)
  ) +
    ggplot2::geom_boxplot(
      ggplot2::aes(fill = .data$B_cell_group),
      width = 0.4
    ) +
    ggpubr::stat_compare_means(
      comparisons = list(c("less", "more")),
      method = "wilcox.test"
    ) +
    ggplot2::theme_bw(base_size = 12) +
    ggplot2::theme(
      legend.position = "none",
      aspect.ratio = 1,
      legend.title = ggplot2::element_blank(),
      axis.title.x = ggplot2::element_blank(),
      plot.title = ggplot2::element_text(
        hjust = 0.5,
        face = "italic",
        size = 9
      ),
      axis.text = ggplot2::element_text(family = "ArialMT"),
      axis.title = ggplot2::element_text(family = "ArialMT"),
      panel.grid = ggplot2::element_blank(),
      panel.border = ggplot2::element_rect(fill = NA, colour = "grey87")
    ) +
    ggplot2::labs(
      y = "Relative abundance",
      title = sub("^s__", "", species_name)
    ) +
    ggplot2::scale_fill_manual(values = group_colors) +
    ggplot2::scale_y_continuous(
      expand = ggplot2::expansion(mult = c(0.05, 0.18))
    )
}

p_blautia_wexlerae <- make_species_boxplot("s__Blautia_wexlerae")
p_blautia_caecimuris <- make_species_boxplot("s__Blautia_caecimuris")
p_ruminococcus_bromii <- make_species_boxplot("s__Ruminococcus_bromii")
p_romboutsia_ilealis <- make_species_boxplot("s__Romboutsia_ilealis")
```

```{r plot-blautia-boxplots, fig.width=5.69, fig.height=2.66}
p_blautia_boxplots <- ggpubr::ggarrange(
  p_blautia_wexlerae,
  p_blautia_caecimuris,
  align = "hv",
  ncol = 2
)

ggplot2::ggsave(
  file.path(figure_dir, "good_B_cell_boxplots_blautia.pdf"),
  p_blautia_boxplots,
  width = 5.69,
  height = 2.66
)

p_blautia_boxplots
```

```{r plot-ruminococcus-romboutsia-boxplots, fig.width=5.69, fig.height=2.66}
p_ruminococcus_romboutsia_boxplots <- ggpubr::ggarrange(
  p_ruminococcus_bromii,
  p_romboutsia_ilealis,
  align = "hv",
  ncol = 2
)

ggplot2::ggsave(
  file.path(figure_dir, "good_B_cell_boxplots_ruminococcus_romboutsia.pdf"),
  p_ruminococcus_romboutsia_boxplots,
  width = 5.69,
  height = 2.66
)

p_ruminococcus_romboutsia_boxplots
```

# 5. Session information

```{r session-information}
sessionInfo()
```
