---
title: "Baseline bacterial score models for M1 outcome prediction"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: true
---

This analysis trains baseline gut bacterial score models for predicting the M1
response class and evaluates each fixed baseline model at D0 and D14. Three
predefined feature configurations are assessed: D14-associated bacteria,
B-cell reconstitution-associated bacteria, and their union.

```{r setup, include=FALSE}
options(stringsAsFactors = FALSE)

suppressPackageStartupMessages({
  library(dplyr)
  library(openxlsx)
  library(ggplot2)
})

input_file <- "000.input16S.Rds"
d14_feature_file <- "000.D14_associated_bacteria.csv"
bcell_feature_file <- "000.B_cell_reconstitution_associated_bacteria.csv"

analysis_name <- "07.outcomeM1_AUC"
output_root <- analysis_name
dir.create(output_root, recursive = TRUE, showWarnings = FALSE)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = file.path(output_root, "knitr_figures", "")
)
```

# 1. Adjustable parameters and model definitions

```{r parameters}
train_time <- "001_BL"
valid_times <- c("002_D0", "003_D14")
all_times <- c(train_time, valid_times)

score_high_class <- "early"
score_other_class <- setdiff(c("early", "later"), score_high_class)
positive_class <- score_high_class
random_seed <- 20260712
bootstrap_iterations <- 2000
permutation_iterations <- 5000

# Select any subset of these model IDs when a partial rerun is needed.
models_to_run <- c(
  "D14_associated",
  "B_cell_reconstitution",
  "Union_set"
)
```

# 2. Input preparation

```{r prepare-input}
required_files <- c(input_file, d14_feature_file, bcell_feature_file)
missing_files <- required_files[!file.exists(required_files)]
if (length(missing_files) > 0L) {
  stop("Missing input file(s): ", paste(missing_files, collapse = ", "))
}

input16S <- readRDS(input_file)
if (!is.list(input16S) || !all(c("dat_rela", "metainfo") %in% names(input16S))) {
  stop("000.input16S.Rds must contain dat_rela and metainfo.")
}

dat_rela <- input16S$dat_rela
metainfo <- input16S$metainfo
d14_features <- utils::read.csv(d14_feature_file, check.names = FALSE)
bcell_features <- utils::read.csv(bcell_feature_file, check.names = FALSE)

required_metadata <- c("Sample_Name", "timeIDcontiue", "outcome_M1")
missing_metadata <- setdiff(required_metadata, names(metainfo))
if (length(missing_metadata) > 0L) {
  stop("Missing metadata column(s): ", paste(missing_metadata, collapse = ", "))
}
if (!"Names" %in% names(d14_features) || !"Names" %in% names(bcell_features)) {
  stop("Each feature input must contain a Names column.")
}
if (!identical(colnames(dat_rela), as.character(metainfo$Sample_Name))) {
  stop("dat_rela columns are not aligned with metainfo$Sample_Name.")
}

metadata_select <- metainfo |>
  dplyr::filter(.data$timeIDcontiue %in% all_times) |>
  dplyr::mutate(
    Sample_Name = as.character(.data$Sample_Name),
    ResponseM1 = dplyr::case_when(
      .data$outcome_M1 == "Fast_responders" ~ "early",
      .data$outcome_M1 == "Delayed_responders" ~ "later",
      TRUE ~ NA_character_
    ),
    ResponseM1 = factor(.data$ResponseM1, levels = c("early", "later"))
  ) |>
  dplyr::filter(!is.na(.data$ResponseM1))

d14_bacteria <- intersect(unique(d14_features$Names), rownames(dat_rela))
bcell_bacteria <- intersect(unique(bcell_features$Names), rownames(dat_rela))
union_bacteria <- intersect(
  unique(c(d14_features$Names, bcell_features$Names)),
  rownames(dat_rela)
)

model_specifications <- list(
  D14_associated = list(
    folder = "01_D14_associated_score_model",
    display_name = "D14-associated score model",
    feature_source = "D14-associated LEfSe bacteria",
    bacteria = d14_bacteria
  ),
  B_cell_reconstitution = list(
    folder = "02_B_cell_reconstitution_associated_score_model",
    display_name = "B-cell reconstitution-associated score model",
    feature_source = "B-cell reconstitution-associated LEfSe bacteria",
    bacteria = bcell_bacteria
  ),
  Union_set = list(
    folder = "03_Union_set_score_model",
    display_name = "Union-set score model",
    feature_source = "Union of D14-associated and B-cell reconstitution-associated bacteria",
    bacteria = union_bacteria
  )
)

unknown_model_ids <- setdiff(models_to_run, names(model_specifications))
if (length(unknown_model_ids) > 0L) {
  stop("Unknown model ID(s): ", paste(unknown_model_ids, collapse = ", "))
}

feature_summary <- data.frame(
  model_id = names(model_specifications),
  input_features = c(
    length(d14_bacteria),
    length(bcell_bacteria),
    length(union_bacteria)
  )
)

sample_summary <- metadata_select |>
  dplyr::count(.data$timeIDcontiue, .data$ResponseM1, name = "n")

feature_summary
sample_summary
```

# 3. Modeling utilities

```{r modeling-functions}
as_numeric_matrix <- function(data) {
  data <- as.matrix(data)
  storage.mode(data) <- "numeric"
  data
}

get_bacteria_pseudocount <- function(data) {
  nonzero <- data[is.finite(data) & data > 0]
  ifelse(length(nonzero) > 0L, min(nonzero) / 2, 1e-6)
}

get_timepoint_data <- function(timepoint, bacteria) {
  metadata_timepoint <- metadata_select |>
    dplyr::filter(.data$timeIDcontiue == timepoint)

  sample_ids <- metadata_timepoint$Sample_Name
  bacteria_raw <- t(
    as_numeric_matrix(
      dat_rela[bacteria, sample_ids, drop = FALSE]
    )
  )
  rownames(bacteria_raw) <- sample_ids

  list(meta = metadata_timepoint, bacteria_raw = bacteria_raw)
}

get_direction <- function(train_z, y_train, high_class = score_high_class) {
  valid_classes <- c("early", "later")
  if (!high_class %in% valid_classes) {
    stop("score_high_class must be either 'early' or 'later'.")
  }
  other_class <- setdiff(valid_classes, high_class)

  median_high <- apply(
    train_z[y_train == high_class, , drop = FALSE],
    2,
    stats::median,
    na.rm = TRUE
  )
  median_other <- apply(
    train_z[y_train == other_class, , drop = FALSE],
    2,
    stats::median,
    na.rm = TRUE
  )
  direction <- sign(median_high - median_other)
  direction[!is.finite(direction)] <- 0
  direction
}

fit_bacteria_score_model <- function(train_x, y_train) {
  keep_available <- apply(
    train_x,
    2,
    function(values) {
      sum(!is.na(values)) >= 4L &&
        length(unique(values[!is.na(values)])) > 1L
    }
  )
  train_x <- train_x[, keep_available, drop = FALSE]

  center <- apply(train_x, 2, stats::median, na.rm = TRUE)
  spread <- apply(train_x, 2, stats::IQR, na.rm = TRUE)
  keep_variable <- is.finite(center) & is.finite(spread) & spread > 0
  train_x <- train_x[, keep_variable, drop = FALSE]
  center <- center[keep_variable]
  spread <- spread[keep_variable]

  train_z <- sweep(sweep(train_x, 2, center, "-"), 2, spread, "/")
  direction <- get_direction(train_z, y_train)
  keep_direction <- direction != 0

  if (!any(keep_direction)) {
    stop("No usable model features remain after filtering and direction assignment.")
  }

  list(
    score_type = "bacteria",
    features = colnames(train_z)[keep_direction],
    center = center[keep_direction],
    spread = spread[keep_direction],
    direction = direction[keep_direction],
    n_feature = sum(keep_direction)
  )
}

predict_bacteria_score <- function(model, new_x) {
  missing_features <- setdiff(model$features, colnames(new_x))
  if (length(missing_features) > 0L) {
    added <- matrix(
      NA_real_,
      nrow = nrow(new_x),
      ncol = length(missing_features),
      dimnames = list(rownames(new_x), missing_features)
    )
    new_x <- cbind(new_x, added)
  }
  new_x <- new_x[, model$features, drop = FALSE]
  new_z <- sweep(sweep(new_x, 2, model$center, "-"), 2, model$spread, "/")
  rowMeans(sweep(new_z, 2, model$direction, "*"), na.rm = TRUE)
}

calculate_auc <- function(score, outcome, positive = positive_class) {
  complete <- !is.na(score) & !is.na(outcome)
  score <- score[complete]
  outcome <- outcome[complete]
  positive_score <- score[outcome == positive]
  negative_score <- score[outcome != positive]
  if (length(positive_score) < 1L || length(negative_score) < 1L) {
    return(NA_real_)
  }
  comparison <- outer(positive_score, negative_score, "-")
  (
    sum(comparison > 0) + 0.5 * sum(comparison == 0)
  ) / (length(positive_score) * length(negative_score))
}

auc_bootstrap_ci <- function(score, outcome, B = bootstrap_iterations) {
  complete <- !is.na(score) & !is.na(outcome)
  score <- score[complete]
  outcome <- outcome[complete]
  positive_ids <- which(outcome == positive_class)
  negative_ids <- which(outcome != positive_class)
  if (length(positive_ids) < 1L || length(negative_ids) < 1L) {
    return(c(NA_real_, NA_real_))
  }

  bootstrap_auc <- replicate(B, {
    sampled_ids <- c(
      sample(positive_ids, length(positive_ids), replace = TRUE),
      sample(negative_ids, length(negative_ids), replace = TRUE)
    )
    calculate_auc(
      score[sampled_ids],
      outcome[sampled_ids],
      positive = positive_class
    )
  })
  stats::quantile(bootstrap_auc, c(0.025, 0.975), na.rm = TRUE)
}

auc_permutation_p <- function(score, outcome, B = permutation_iterations) {
  observed <- calculate_auc(score, outcome, positive = positive_class)
  if (is.na(observed)) {
    return(NA_real_)
  }
  null_auc <- replicate(
    B,
    calculate_auc(score, sample(outcome), positive = positive_class)
  )
  (
    sum(abs(null_auc - 0.5) >= abs(observed - 0.5), na.rm = TRUE) + 1
  ) / (B + 1)
}

roc_points <- function(score, outcome, score_name, timepoint) {
  complete <- !is.na(score) & !is.na(outcome)
  score <- score[complete]
  outcome <- outcome[complete]
  thresholds <- c(Inf, sort(unique(score), decreasing = TRUE), -Inf)

  dplyr::bind_rows(lapply(thresholds, function(threshold) {
    prediction <- score >= threshold
    true_positive <- sum(prediction & outcome == positive_class)
    false_positive <- sum(prediction & outcome != positive_class)
    false_negative <- sum(!prediction & outcome == positive_class)
    true_negative <- sum(!prediction & outcome != positive_class)

    data.frame(
      threshold = threshold,
      sensitivity = ifelse(
        true_positive + false_negative > 0,
        true_positive / (true_positive + false_negative),
        NA_real_
      ),
      specificity = ifelse(
        true_negative + false_positive > 0,
        true_negative / (true_negative + false_positive),
        NA_real_
      ),
      score_type = score_name,
      timepoint = timepoint
    )
  }))
}

make_auc_row <- function(timepoint, score, outcome, validation_type, n_feature) {
  confidence_interval <- auc_bootstrap_ci(score, outcome)
  data.frame(
    score_type = "bacteria",
    timepoint = timepoint,
    validation_type = validation_type,
    n_total = sum(!is.na(score) & !is.na(outcome)),
    n_early = sum(!is.na(score) & outcome == "early"),
    n_later = sum(!is.na(score) & outcome == "later"),
    n_feature = n_feature,
    AUC = calculate_auc(score, outcome, positive = positive_class),
    CI_low = as.numeric(confidence_interval[1]),
    CI_high = as.numeric(confidence_interval[2]),
    permutation_P = auc_permutation_p(score, outcome)
  )
}

format_p <- function(p_value) {
  ifelse(
    is.na(p_value),
    "P=NA",
    ifelse(p_value < 0.001, "P<0.001", sprintf("P=%.3f", p_value))
  )
}
```

# 4. Run one model configuration

```{r run-model-function}
run_score_model <- function(model_id, specification) {
  # Resetting the original seed makes each configuration reproducible as a
  # standalone run of the source workflow.
  set.seed(random_seed)

  bacteria <- specification$bacteria
  if (length(bacteria) == 0L) {
    stop("No input bacteria remain for model: ", model_id)
  }

  model_dir <- file.path(output_root, specification$folder)
  figure_dir <- file.path(model_dir, "figures")
  dir.create(model_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

  train_data <- get_timepoint_data(train_time, bacteria)
  y_train <- train_data$meta$ResponseM1
  bacteria_pseudocount_M0 <- get_bacteria_pseudocount(train_data$bacteria_raw)
  train_x <- log10(train_data$bacteria_raw + bacteria_pseudocount_M0)

  baseline_model <- fit_bacteria_score_model(train_x, y_train)
  train_score_apparent <- predict_bacteria_score(baseline_model, train_x)

  train_score_LOOCV <- rep(NA_real_, length(y_train))
  n_feature_LOOCV <- rep(NA_integer_, length(y_train))

  for (index in seq_along(y_train)) {
    training_ids <- setdiff(seq_along(y_train), index)
    raw_training <- train_data$bacteria_raw[training_ids, , drop = FALSE]
    raw_testing <- train_data$bacteria_raw[index, , drop = FALSE]
    pseudocount_fold <- get_bacteria_pseudocount(raw_training)
    x_training <- log10(raw_training + pseudocount_fold)
    x_testing <- log10(raw_testing + pseudocount_fold)
    model_fold <- fit_bacteria_score_model(x_training, y_train[training_ids])
    train_score_LOOCV[index] <- predict_bacteria_score(model_fold, x_testing)
    n_feature_LOOCV[index] <- model_fold$n_feature
  }

  score_list <- list()
  auc_list <- list()
  sample_count_list <- list()
  roc_list <- list()

  score_list[[train_time]] <- data.frame(
    timepoint = train_time,
    Sample_Name = train_data$meta$Sample_Name,
    ResponseM1 = y_train,
    bacteria_score_M0_model = train_score_apparent,
    bacteria_score_M0_LOOCV = train_score_LOOCV
  )
  auc_list[[paste0(train_time, "_apparent")]] <- make_auc_row(
    train_time,
    train_score_apparent,
    y_train,
    "M0_apparent_fixed_model",
    baseline_model$n_feature
  )
  auc_list[[paste0(train_time, "_LOOCV")]] <- make_auc_row(
    train_time,
    train_score_LOOCV,
    y_train,
    "M0_LOOCV_internal",
    stats::median(n_feature_LOOCV, na.rm = TRUE)
  )
  sample_count_list[[train_time]] <- data.frame(
    timepoint = train_time,
    n_included = nrow(train_data$meta),
    n_early = sum(y_train == "early"),
    n_later = sum(y_train == "later")
  )
  roc_list[[paste0(train_time, "_apparent")]] <- roc_points(
    train_score_apparent,
    y_train,
    "bacteria_apparent",
    train_time
  )
  roc_list[[paste0(train_time, "_LOOCV")]] <- roc_points(
    train_score_LOOCV,
    y_train,
    "bacteria_LOOCV",
    train_time
  )

  for (timepoint in valid_times) {
    validation_data <- get_timepoint_data(timepoint, bacteria)
    y_validation <- validation_data$meta$ResponseM1
    x_validation <- log10(
      validation_data$bacteria_raw + bacteria_pseudocount_M0
    )
    validation_score <- predict_bacteria_score(
      baseline_model,
      x_validation
    )

    score_list[[timepoint]] <- data.frame(
      timepoint = timepoint,
      Sample_Name = validation_data$meta$Sample_Name,
      ResponseM1 = y_validation,
      bacteria_score_M0_model = validation_score,
      bacteria_score_M0_LOOCV = NA_real_
    )
    auc_list[[timepoint]] <- make_auc_row(
      timepoint,
      validation_score,
      y_validation,
      "external_timepoint_fixed_M0_model",
      baseline_model$n_feature
    )
    sample_count_list[[timepoint]] <- data.frame(
      timepoint = timepoint,
      n_included = nrow(validation_data$meta),
      n_early = sum(y_validation == "early"),
      n_later = sum(y_validation == "later")
    )
    roc_list[[timepoint]] <- roc_points(
      validation_score,
      y_validation,
      "bacteria",
      timepoint
    )
  }

  scores_all <- dplyr::bind_rows(score_list)
  auc_summary <- dplyr::bind_rows(auc_list)
  sample_counts <- dplyr::bind_rows(sample_count_list)
  roc_data <- dplyr::bind_rows(roc_list)

  auc_label_data <- auc_summary |>
    dplyr::mutate(
      score_type = dplyr::case_when(
        .data$validation_type == "M0_apparent_fixed_model" ~ "bacteria_apparent",
        .data$validation_type == "M0_LOOCV_internal" ~ "bacteria_LOOCV",
        TRUE ~ "bacteria"
      ),
      label_y = dplyr::case_when(
        .data$validation_type == "M0_apparent_fixed_model" ~ 0.26,
        .data$validation_type == "M0_LOOCV_internal" ~ 0.06,
        TRUE ~ 0.06
      ),
      label = sprintf(
        "AUC=%.3f\n95%% CI %.3f-%.3f\n%s",
        .data$AUC,
        .data$CI_low,
        .data$CI_high,
        format_p(.data$permutation_P)
      )
    )

  auc_ci_plot_data <- auc_summary |>
    dplyr::mutate(
      analysis_label = dplyr::case_when(
        .data$validation_type == "M0_apparent_fixed_model" ~ "M0 apparent",
        .data$validation_type == "M0_LOOCV_internal" ~ "M0 LOOCV",
        .data$timepoint == "002_D0" ~ "D0 validation",
        .data$timepoint == "003_D14" ~ "D14 validation",
        TRUE ~ .data$timepoint
      ),
      analysis_label = factor(
        .data$analysis_label,
        levels = c(
          "M0 apparent",
          "M0 LOOCV",
          "D0 validation",
          "D14 validation"
        )
      ),
      label = sprintf(
        "AUC %.3f, %s",
        .data$AUC,
        format_p(.data$permutation_P)
      )
    )

  feature_direction <- data.frame(
    score_type = "bacteria",
    feature = baseline_model$features,
    center_M0 = baseline_model$center,
    IQR_M0 = baseline_model$spread,
    direction = baseline_model$direction,
    direction_label = ifelse(
      baseline_model$direction > 0,
      paste0("higher_in_", score_high_class),
      paste0("higher_in_", score_other_class)
    ),
    stringsAsFactors = FALSE
  )

  model_information <- list(
    model_id = model_id,
    model_display_name = specification$display_name,
    train_time = train_time,
    valid_times = valid_times,
    score_high_class = score_high_class,
    positive_class = positive_class,
    random_seed = random_seed,
    bootstrap_iterations = bootstrap_iterations,
    permutation_iterations = permutation_iterations,
    bacteria_pseudocount_M0 = bacteria_pseudocount_M0,
    bacteria_features_input = bacteria,
    model = baseline_model,
    note = paste(
      "Only bacteria are used. D0 and D14 use the fixed M0/BL feature set,",
      "median/IQR scaling and direction."
    )
  )
  saveRDS(
    model_information,
    file.path(model_dir, "M0_bacteria_ResponseM1_score_model.rds")
  )

  notes <- data.frame(
    item = c(
      "model_configuration",
      "training_timepoint",
      "validation_timepoints",
      "feature_source",
      "score_definition",
      "score_high_class",
      "positive_class",
      "important_caution"
    ),
    description = c(
      specification$display_name,
      "M0 is coded as 001_BL in the metadata.",
      "D0 and D14 are scored using the fixed M0 bacteria model.",
      specification$feature_source,
      paste(
        "Bacteria abundance is log10(relative abundance + M0 pseudocount),",
        paste0(
          "standardized by M0 median/IQR, signed by M0 ",
          score_high_class,
          "-vs-",
          score_other_class,
          " direction,"
        ),
        "then averaged."
      ),
      paste0("Higher scores indicate a more ", score_high_class, "-like profile."),
      paste0(
        "AUC treats ",
        positive_class,
        " as the positive class, consistent with the score direction."
      ),
      paste(
        "D0 and D14 are temporal validation timepoints from the same cohort,",
        "not independent external validation."
      )
    )
  )

  utils::write.csv(
    notes,
    file.path(model_dir, "00_model_notes.csv"),
    row.names = FALSE,
    fileEncoding = "UTF-8"
  )
  utils::write.csv(
    scores_all,
    file.path(model_dir, "01_scores_in_M0_D0_D14.csv"),
    row.names = FALSE,
    fileEncoding = "UTF-8"
  )
  utils::write.csv(
    auc_summary,
    file.path(model_dir, "02_AUC_in_M0_D0_D14.csv"),
    row.names = FALSE,
    fileEncoding = "UTF-8"
  )
  utils::write.csv(
    feature_direction,
    file.path(model_dir, "03_feature_direction.csv"),
    row.names = FALSE,
    fileEncoding = "UTF-8"
  )
  utils::write.csv(
    sample_counts,
    file.path(model_dir, "04_sample_counts.csv"),
    row.names = FALSE,
    fileEncoding = "UTF-8"
  )

  openxlsx::write.xlsx(
    list(
      notes = notes,
      AUC_summary = auc_summary,
      scores = scores_all,
      feature_direction = feature_direction,
      sample_counts = sample_counts
    ),
    file.path(model_dir, "M0_bacteria_model_ResponseM1_AUC_summary.xlsx"),
    overwrite = TRUE
  )

  plot_title_prefix <- specification$display_name

  roc_plot <- ggplot2::ggplot(
    roc_data,
    ggplot2::aes(
      x = 1 - .data$specificity,
      y = .data$sensitivity,
      color = .data$score_type
    )
  ) +
    ggplot2::geom_abline(
      slope = 1,
      intercept = 0,
      linetype = "dashed",
      color = "grey60"
    ) +
    ggplot2::geom_step(linewidth = 0.8) +
    ggplot2::geom_label(
      data = auc_label_data,
      ggplot2::aes(
        x = 0.03,
        y = .data$label_y,
        label = .data$label,
        color = .data$score_type
      ),
      inherit.aes = FALSE,
      hjust = 0,
      vjust = 0,
      size = 2.6,
      linewidth = 0.2,
      fill = "white",
      alpha = 0.9,
      show.legend = FALSE
    ) +
    ggplot2::coord_equal() +
    ggplot2::facet_wrap(~timepoint) +
    ggplot2::theme_bw(base_size = 11) +
    ggplot2::labs(
      x = "1 - specificity",
      y = "sensitivity",
      color = "score",
      title = paste0(plot_title_prefix, ": ResponseM1 ROC")
    )

  ggplot2::ggsave(
    file.path(figure_dir, "Figure_01_ROC_in_M0_D0_D14.pdf"),
    roc_plot,
    width = 8,
    height = 4.5
  )

  auc_ci_plot <- ggplot2::ggplot(
    auc_ci_plot_data,
    ggplot2::aes(x = .data$analysis_label, y = .data$AUC)
  ) +
    ggplot2::geom_hline(
      yintercept = 0.5,
      linetype = "dashed",
      color = "grey60"
    ) +
    ggplot2::geom_pointrange(
      ggplot2::aes(ymin = .data$CI_low, ymax = .data$CI_high),
      linewidth = 0.7,
      size = 0.8,
      color = "#2563EB"
    ) +
    ggplot2::geom_text(
      ggplot2::aes(label = .data$label),
      hjust = -0.05,
      size = 3.2
    ) +
    ggplot2::coord_flip(ylim = c(0, 1.08), clip = "off") +
    ggplot2::theme_bw(base_size = 11) +
    ggplot2::theme(
      plot.margin = ggplot2::margin(5.5, 50, 5.5, 5.5)
    ) +
    ggplot2::labs(
      x = NULL,
      y = "AUC with bootstrap 95% CI",
      title = paste0(plot_title_prefix, ": AUC")
    )

  ggplot2::ggsave(
    file.path(figure_dir, "Figure_02_AUC_CI_in_M0_D0_D14.pdf"),
    auc_ci_plot,
    width = 7,
    height = 4
  )

  score_plot_data <- scores_all |>
    dplyr::mutate(
      plot_score = ifelse(
        .data$timepoint == train_time,
        .data$bacteria_score_M0_LOOCV,
        .data$bacteria_score_M0_model
      ),
      score_version = ifelse(
        .data$timepoint == train_time,
        "M0_LOOCV",
        "fixed_M0_model"
      )
    )

  score_boxplot <- ggplot2::ggplot(
    score_plot_data,
    ggplot2::aes(
      x = .data$ResponseM1,
      y = .data$plot_score,
      fill = .data$ResponseM1
    )
  ) +
    ggplot2::geom_boxplot(
      width = 0.55,
      outlier.shape = NA,
      alpha = 0.65
    ) +
    ggplot2::geom_jitter(width = 0.12, size = 1.8, alpha = 0.85) +
    ggplot2::facet_wrap(~timepoint, scales = "free_y") +
    ggplot2::theme_bw(base_size = 11) +
    ggplot2::theme(legend.position = "none") +
    ggplot2::labs(
      x = NULL,
      y = "Bacteria score",
      title = paste0(plot_title_prefix, ": scores in M0, D0 and D14")
    )

  ggplot2::ggsave(
    file.path(figure_dir, "Figure_03_score_boxplot_in_M0_D0_D14.pdf"),
    score_boxplot,
    width = 8,
    height = 4.5
  )

  list(
    model_id = model_id,
    display_name = specification$display_name,
    model_dir = model_dir,
    input_feature_count = length(bacteria),
    fitted_feature_count = baseline_model$n_feature,
    AUC_summary = auc_summary,
    sample_counts = sample_counts,
    ROC_plot = roc_plot,
    AUC_CI_plot = auc_ci_plot,
    score_boxplot = score_boxplot
  )
}
```

# 5. Run the selected model configurations

```{r run-all-models}
model_results <- lapply(models_to_run, function(model_id) {
  run_score_model(model_id, model_specifications[[model_id]])
})
names(model_results) <- models_to_run

combined_auc_summary <- dplyr::bind_rows(lapply(
  model_results,
  function(result) {
    result$AUC_summary |>
      dplyr::mutate(
        model_id = result$model_id,
        model_name = result$display_name,
        input_feature_count = result$input_feature_count,
        fitted_feature_count = result$fitted_feature_count,
        .before = 1
      )
  }
))

utils::write.csv(
  combined_auc_summary,
  file.path(output_root, "00_three_model_AUC_summary.csv"),
  row.names = FALSE,
  fileEncoding = "UTF-8"
)

combined_auc_summary
```

# 6. Model figures

## 6.1 D14-associated score model

```{r show-d14-model-figures, fig.width=8, fig.height=4.5}
if ("D14_associated" %in% names(model_results)) {
  print(model_results$D14_associated$ROC_plot)
  print(model_results$D14_associated$AUC_CI_plot)
  print(model_results$D14_associated$score_boxplot)
}
```

## 6.2 B-cell reconstitution-associated score model

```{r show-bcell-model-figures, fig.width=8, fig.height=4.5}
if ("B_cell_reconstitution" %in% names(model_results)) {
  print(model_results$B_cell_reconstitution$ROC_plot)
  print(model_results$B_cell_reconstitution$AUC_CI_plot)
  print(model_results$B_cell_reconstitution$score_boxplot)
}
```

## 6.3 Union-set score model

```{r show-union-model-figures, fig.width=8, fig.height=4.5}
if ("Union_set" %in% names(model_results)) {
  print(model_results$Union_set$ROC_plot)
  print(model_results$Union_set$AUC_CI_plot)
  print(model_results$Union_set$score_boxplot)
}
```

# 7. Session information

```{r session-information}
sessionInfo()
```
